---
title: "01_input_preparation"
author: "Jordan H. Whitlock & Tabea M. Soelter"
date: "2023-11-07"
output: html_document
---
**Creating input for files for gene regulatory network generation with PANDA**

__Code Author__: JHW performed these analyses. TMS adjusted the code upon re-generation and to increase reproducibility.

__Goal__: Splitting pseudo bulk files by time point and creating input text files for downstream analysis.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*Pseudo bulk matrices*
* Names:
  1. excitatory_neurons.rds
  2. inhibitory_neurons.rds
* Location: data/pseudobulk/
*Sample sheet*
* Name: sample_sheet.rds
* Location: doc/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * pseudo bulk matrices
  * sample sheet
* Split pseudo bulk matrices by time point
  * Save split files
* Create text file with file paths to split pseudo bulk matrices
  * Save text file
  
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(purrr)
  library(Matrix)
  library(tidyverse)
  library(lintr)
  library(styler)
  library(here)
})
```

# Load data
* Loading and adding to list 
* Only including receiver cell types
```{r}
# Get file paths to neuronal pseudo bulk files
file_paths <- list.files(here("data", "pseudobulk"),
  pattern = "neurons",
  full.names = TRUE
)

# Load files into list
expression_list <- map(file_paths, ~ readRDS(.x))

# Load sample sheet
sample_sheet <- readRDS(here("doc/sample_sheet.rds"))
```

# Create necessary data directories
```{r}
directories <- c(
  here("data/pseudobulk_split"),
  here("data/PANDA_inputs")
)

lapply(directories[!directories %in% dir()], dir.create)
```

# Convert pseudo bulk matrices to df
```{r}
# Create a new list to convert the dgCMatrix to data frames
data_frame_list <- list()

# Convert each dgCMatrix to a data frame and store in data_frame_list
for (matrix in expression_list) {
  matrix_data <- as(matrix, "matrix")
  data_frame <- as.data.frame(matrix_data)
  data_frame_list <- append(data_frame_list, list(data_frame))
}

# Setting the names of the list elements to match the variable names
names(data_frame_list) <- c("excitatory_neurons", "inhibitory_neurons")

# Fixing the column names to remove everything before the sample name
data_frame_list <- lapply(data_frame_list, function(df) {
  colnames(df) <- sub(".*_S", "S", colnames(df))
  df
})
```

# Split matrices by time point
* We split pseudo bulk matrices using info from the sample sheet.
* Matrices include both conditions and time points for every cell type, but we need group-specific matrices
```{r}
# Create a column with time point and condition info on sample sheet
sample_sheet <- sample_sheet %>%
  unite(group, c(orig.ident, timepoint), sep = "", remove = FALSE)

sample_sheet$group <- gsub(sample_sheet$group,
  pattern = "m",
  replacement = ""
)

# Grab group vector
groups <- as.vector(sample_sheet$group) %>% unique()

# Grab sample vector for each group
samples <- lapply(split(sample_sheet$sample, sample_sheet$group), as.vector) %>%
  as.list()

# Setting the output directory for split matrices
common_path <- here("data/pseudobulk_split/")

# Creating a list to store the new data frames
timepoint_expression_list <- list()

# Loop through the list of data frames
for (object_name in names(data_frame_list)) {
  current_df <- data_frame_list[[object_name]]

  for (time_point in groups) {
    current_columns <- samples[[time_point]]
    new_df <- current_df[, current_columns, drop = FALSE]
    new_name <- paste0(object_name, "_", time_point)
    timepoint_expression_list[[new_name]] <- new_df
    save(new_df, file = file.path(common_path, paste0(new_name, ".Rdata")))
  }
}
```

# Generating PANDA array input text file
```{r}
files <- list.files(here("data", "pseudobulk_split"), full.names = TRUE)

write.table(files,
  file = here("data/PANDA_inputs/PANDA_exp_files_array.txt"),
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)
```

# Session Info
```{r}
sessionInfo()
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] here_1.0.1      styler_1.9.1    lintr_3.0.2     lubridate_1.9.2 forcats_1.0.0  
 [6] stringr_1.5.0   dplyr_1.1.1     readr_2.1.4     tidyr_1.3.0     tibble_3.2.1   
[11] ggplot2_3.4.2   tidyverse_2.0.0 Matrix_1.5-4    purrr_1.0.1    

loaded via a namespace (and not attached):
 [1] tidyselect_1.2.0  xfun_0.38         remotes_2.4.2     lattice_0.21-8   
 [5] colorspace_2.1-0  vctrs_0.6.2       generics_0.1.3    htmltools_0.5.5  
 [9] yaml_2.3.7        utf8_1.2.3        rlang_1.1.0       R.oo_1.25.0      
[13] pillar_1.9.0      R.utils_2.12.2    glue_1.6.2        withr_2.5.0      
[17] lifecycle_1.0.3   R.cache_0.16.0    munsell_0.5.0     gtable_0.3.3     
[21] R.methodsS3_1.8.2 evaluate_0.20     knitr_1.42        tzdb_0.3.0       
[25] callr_3.7.3       fastmap_1.1.1     ps_1.7.5          rex_1.2.1        
[29] fansi_1.0.4       scales_1.2.1      desc_1.4.2        hms_1.1.3        
[33] digest_0.6.31     stringi_1.7.12    processx_3.8.1    grid_4.2.3       
[37] rprojroot_2.0.3   cli_3.6.1         tools_4.2.3       magrittr_2.0.3   
[41] lazyeval_0.2.2    crayon_1.5.2      pkgconfig_2.0.3   xml2_1.3.3       
[45] cyclocomp_1.1.0   timechange_0.2.0  rmarkdown_2.21    rstudioapi_0.14  
[49] R6_2.5.1          compiler_4.2.3

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "gene_targeting",
  "01_input_preparation.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "gene_targeting",
  "01_input_preparation.Rmd"
))
```
