---
title: "03_gene_targeting"
author: "Tabea M. Soelter & Jordan H. Whitlock"
date: "2024-02-11"
output: html_document
---
**Calculate differential gene targeting of signaling mediators**

__Goal__: Determine differential gene targeting scores within receivers for signaling mediators between conditions from PANDA regulatory networks.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.3
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*PANDA GRNs*
* Names:
  1. excitatory_neurons_AD12.Rdata
  2. excitatory_neurons_WT12.Rdata
  3. inhibitory_neurons_AD12.Rdata
  4. inhibitory_neurons_WT12.Rdata
* Location: data/panda/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * PANDA networks
* Calculate gene targeting score
* Calculate differential gene targeting score
* Determine quartile cut-offs and filter
* Save objects
  * AD enriched (unfiltered)
  * AD enriched (filtered)
  
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(pandaR)
  library(tidyverse)
  library(tibble)
  library(data.table)
  library(limma)
  library(magrittr)
  library(readr)
  library(gprofiler2)
  library(ggridges)
  library(lintr)
  library(here)
  library(styler)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
files <- list.files(here("data", "panda"),
  pattern = "neurons.+12",
  full.names = TRUE
)

for (i in files) {
  load(i)
  reg_net <- pandaResults
  reg_net <- reg_net@regNet
  name <- sub(here("data/panda/"), "", i)
  name <- sub(".Rdata*", "", name)
  assign(paste0(name, ".regNet"), reg_net)
  rm(pandaResults)
  rm(reg_net)
}

# Create vector with object names
networks <- basename(files)
networks <- sub(".Rdata.*", "", networks)
```

# Calculate targeting score
```{r}
lapply(networks, calculate_gene_targeting)
```

# Combine AD and WT scores
```{r}
# Grab cell type names (should be excitatory and inhibitory neurons)
cell_names <- networks[grep("_WT", networks)] %>% sub("_WT.*", "", .)

# Combine AD and WT scores into 1 df for each cell type
targeting_list <- list()

for (i in cell_names) {
  x <- get(paste0(i, "_gene_targeting_WT"))
  y <- get(paste0(i, "_gene_targeting_AD"))
  gene_targeting <- merge(x, y, by = "gene")
  targeting_list[[paste0("Gene_targeting_", i)]] <- gene_targeting
}
```

# Construct AD and WT targeting matrices
```{r}
gene_targeting_all_ad <- construct_targeting_mat(
  targeting_list,
  edge_type = "ad_edge_weight_pos"
)

dim(gene_targeting_all_ad) # 19750 3

gene_targeting_all_wt <- construct_targeting_mat(
  targeting_list,
  edge_type = "wt_edge_weight_pos"
)

dim(gene_targeting_all_wt) # 19750 3
```

# Combine all scores
```{r warning=FALSE}
combined <- merge(gene_targeting_all_wt,
  gene_targeting_all_ad,
  by = "gene"
)

# add column names and melt
colnames(combined) <- c(
  "gene",
  "excitatory_neurons_wt",
  "inhibitory_neurons_wt",
  "excitatory_neurons_ad",
  "inhibitory_neurons_ad"
)

combined <- melt(combined)

# add cell type and condition annotation info, then combine with df
variable_info <- t(as.data.frame(strsplit(
  sub("(_)(?=[^_]+$)",
    " ",
    combined$variable,
    perl = TRUE
  ),
  " "
)))

combined <- cbind(combined, variable_info)

# assign column names
colnames(combined) <- c(
  "gene",
  "variable_info",
  "targeting_score",
  "cell_type",
  "condition"
)

dim(combined) # 79000 5
```

# Calculate differential gene targeting score
* Calculate the difference (ad-wt) for the targeting weight where the more positive difference indicates higher gene targeting in AD
```{r warning=FALSE}
combined_gene_targeting <- prepare_diff_targeting(
  gene_targeting_all_ad,
  gene_targeting_all_wt
)

# AD enriched targeting
cell_types <- c("excitatory_neurons", "inhibitory_neurons")

ad_enriched <- calc_diff_targeting(combined_gene_targeting,
  cell_names = cell_types
)
```

# Determine quartiles
* In order to categorize genes as differentially targeted, we calculate quartiles
  * Genes in the top quartile have increased regulation compared to other genes
  * Genes in the bottom quartile have decreased regulation
    * Therefore, they are not "middle of of the pack"
```{r}
# calculate quartiles for each cell type
quartiles <- apply(ad_enriched[, -1], 2, quantile, probs = c(0.25, 0.5, 0.75))

print(quartiles) # see output below
```
    excitatory_neurons inhibitory_neurons
25%           19.70247         -144.94422
50%           67.73835          -70.64913
75%          141.41649          -15.93067

# Filter based on quartiles
* <= Q1 and >= Q3 for each cell type
```{r}
# Excitatory neurons
excitatory_neurons <- ad_enriched[, c(1, 2)]
excitatory_neurons <- subset(excitatory_neurons,
                             excitatory_neurons <= 19.70247 |
                               excitatory_neurons >= 141.41649)
colnames(excitatory_neurons) <- c("gene", "targeting_score")
excitatory_neurons$cell_type <- "excitatory_neurons"

# Inhibitory neurons
inhibitory_neurons <- ad_enriched[, c(1, 3)]
inhibitory_neurons <- subset(inhibitory_neurons,
                             inhibitory_neurons <= -144.94422 |
                               inhibitory_neurons >= -15.93067)
colnames(inhibitory_neurons) <- c("gene", "targeting_score")
inhibitory_neurons$cell_type <- "inhibitory_neurons"

# Combine
filtered_ad <- bind_rows(excitatory_neurons, inhibitory_neurons)

dim(filtered_ad) # 19752 3
```

# Save outputs
```{r}
saveRDS(filtered_ad,
  file = here(
    "results",
    "intermediate_outputs",
    "07_gene_targeting",
    "filtered_targeting_ad.rds"
  )
)

saveRDS(ad_enriched,
  file = here(
    "results",
    "intermediate_outputs",
    "07_gene_targeting",
    "ad_enriched.rds"
  )
)
```

# Session Info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.4 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] styler_1.9.1        here_1.0.1          lintr_3.0.2         ggridges_0.5.4     
 [5] gprofiler2_0.2.1    magrittr_2.0.3      limma_3.54.2        data.table_1.14.8  
 [9] lubridate_1.9.2     forcats_1.0.0       stringr_1.5.0       dplyr_1.1.1        
[13] purrr_1.0.1         readr_2.1.4         tidyr_1.3.0         tibble_3.2.1       
[17] ggplot2_3.4.2       tidyverse_2.0.0     pandaR_1.30.0       Biobase_2.58.0     
[21] BiocGenerics_0.44.0

loaded via a namespace (and not attached):
 [1] Rcpp_1.0.10        lattice_0.21-8     ps_1.7.5           rprojroot_2.0.3   
 [5] digest_0.6.31      utf8_1.2.3         R6_2.5.1           plyr_1.8.8        
 [9] evaluate_0.20      httr_1.4.5         pillar_1.9.0       rlang_1.1.0       
[13] lazyeval_0.2.2     rstudioapi_0.14    callr_3.7.3        hexbin_1.28.3     
[17] R.utils_2.12.2     R.oo_1.25.0        RUnit_0.4.32       rmarkdown_2.21    
[21] desc_1.4.2         htmlwidgets_1.6.2  igraph_1.4.2       munsell_0.5.0     
[25] compiler_4.2.3     xfun_0.38          pkgconfig_2.0.3    htmltools_0.5.5   
[29] tidyselect_1.2.0   matrixStats_0.63.0 reshape_0.8.9      fansi_1.0.4       
[33] viridisLite_0.4.1  crayon_1.5.2       tzdb_0.3.0         withr_2.5.0       
[37] R.methodsS3_1.8.2  grid_4.2.3         jsonlite_1.8.4     gtable_0.3.3      
[41] lifecycle_1.0.3    scales_1.2.1       cli_3.6.1          stringi_1.7.12    
[45] reshape2_1.4.4     remotes_2.4.2      rex_1.2.1          xml2_1.3.3        
[49] generics_0.1.3     vctrs_0.6.2        cyclocomp_1.1.0    tools_4.2.3       
[53] R.cache_0.16.0     glue_1.6.2         hms_1.1.3          yaml_2.3.7        
[57] processx_3.8.1     fastmap_1.1.1      timechange_0.2.0   colorspace_2.1-0  
[61] plotly_4.10.1      knitr_1.42

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 10 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "gene_targeting",
  "03_gene_targeting.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "gene_targeting",
  "03_gene_targeting.Rmd"
))
```
