---
title: "figure_S3"
author: "Tabea M. Soelter"
date: "2024-01-19"
output: html_document
---
**Plotting Main Supplementary Figure 2**

__Goal__: This is a script for generating supp figure 2 of my manuscript.
  
__Reproducibility__: 
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
  * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*Jaccard Index outputs*
* Name: jaccard_results_ligands.rds, jaccard_results_receptors.rds, jaccard_results_targets.rds, jaccard_receptors.rds, jaccard_targets.rds
* Location: results/intermediate_outputs/04_ccc/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * JI calculations dfs
* Plot individual panels
  * A: lollipop plot of JI for receptors and targets in receivers
  * B: corr plot of JI of ligands in senders in inhibitory neurons
  * C: corr plot of JI of receptors by senders in inhibitory neurons
  * D: corr plot of JI of targets by senders in inhibitory neurons
  * E: corr plot of JI of ligands in senders in excitatory neurons
  * F: corr plot of JI of receptors by senders in excitatory neurons
  * G: corr plot of JI of targets by senders in excitatory neurons
* Compile figure
* Save compiled figure and compiled corr plots

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(multinichenetr)
  library(patchwork)
  library(cowplot)
  library(ggcorrplot)
  library(ggpubr)
  library(grid)
  library(ggtext)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
# Jaccard Similarity Index dataframes
jaccard_results_ligands <- readRDS(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "jaccard_ligands_senders.rds"
))

jaccard_results_receptors <- readRDS(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "jaccard_receptors_senders.rds"
))

jaccard_results_targets <- readRDS(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "jaccard_targets_senders.rds"
))

jaccard_receptors <- readRDS(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "jaccard_receptors_receivers.rds"
))

jaccard_targets <- readRDS(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "jaccard_targets_receivers.rds"
))
```

# Colors
```{r}
colors <- c(
  "lightblue",
  "royalblue",
  "darkred",
  "darksalmon",
  "goldenrod1"
)
```

# Set Filtering Criteria
* These are the same as previously
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# vector of groups
group_oi <- c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
```

# Panel S3A
```{r}
# add type column for filtering and plotting purposes before binding dfs
jaccard_receptors$type <- "Receptors"
jaccard_targets$type <- "Targets"

jaccard_all <- rbind(jaccard_receptors, jaccard_targets)

# generate plot
lollipop <- jaccard_all %>%
  ggplot() +
  geom_linerange(
    aes(
      x = reorder(sender, desc(sender)),
      ymin = 0,
      ymax = jaccard_index,
      colour = type
    ),
    position = position_dodge(width = 0.2), size = 1
  ) +
  geom_point(
    aes(
      x = reorder(sender, desc(sender)),
      y = jaccard_index,
      colour = type
    ),
    position = position_dodge(width = 0.2), size = 2
  ) +
  coord_flip() +
  scale_color_manual(
    values = c(
      "Receptors" = "darkgreen",
      "Targets" = "darkseagreen"
    ),
    name = "Type"
  ) +
  theme_bw() +
  labs(x = "Sender", y = "JI") +
  ylim(0, 0.5) +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 14)
  )

lollipop

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "lollipop_JI_final.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
lollipop
dev.off()
```

# Panel S3B
```{r}
# prepare df
jaccard_ligands_df <- data.frame(
  cell_types = names(jaccard_results_ligands),
  jaccard_index = unlist(jaccard_results_ligands),
  stringsAsFactors = FALSE
)

# separate receivers from senders
jaccard_ligands_df <- jaccard_ligands_df %>%
  separate(cell_types,
    into = c(
      "receiver",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

jaccard_ligands_df <- jaccard_ligands_df %>%
  spread(
    key = celltype2,
    value = jaccard_index
  )

# prepare inhibitory neurons
jaccard_ligands_in <- jaccard_ligands_df %>%
  filter(receiver == "Inhibitory Neurons")

jaccard_in_mat <- as.matrix(jaccard_ligands_in[-c(1:2)])
rownames(jaccard_in_mat) <- jaccard_ligands_in[[2]]

# plot figure
panel_b <- ggcorrplot(jaccard_in_mat,
  outline.color = "white",
  lab = TRUE,
  lab_size = 5,
  type = "lower"
) +
  scale_fill_gradientn(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    colors = colors,
    name = "JI"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(angle = 90, size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_blank()
  ) +
  labs(y = "Inhibitory Neurons", x = "") +
  ggtitle("Ligands")

panel_b

# save figure
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_ligands_inhibitory.png"
  ),
  width = 150,
  height = 150,
  units = "mm",
  res = 300
)
panel_b
dev.off()
```

# Panel S3C
```{r}
# prepare df same as before, this time for receptors not ligands
jaccard_receptors_df <- data.frame(
  cell_types = names(jaccard_results_receptors),
  jaccard_index = unlist(jaccard_results_receptors),
  stringsAsFactors = FALSE
)

jaccard_receptors_df <- jaccard_receptors_df %>%
  separate(cell_types,
    into = c(
      "receiver",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

jaccard_receptors_df <- jaccard_receptors_df %>%
  spread(
    key = celltype2,
    value = jaccard_index
  )
# prep inhibitory neuron input
jaccard_receptors_in <- jaccard_receptors_df %>%
  filter(receiver == "Inhibitory Neurons")

jaccard_in_mat <- as.matrix(jaccard_receptors_in[-c(1:2)])
rownames(jaccard_in_mat) <- jaccard_receptors_in[[2]]

# make and plot panel
panel_c <- ggcorrplot(jaccard_in_mat,
  outline.color = "white",
  lab = TRUE,
  lab_size = 5,
  type = "lower"
) +
  scale_fill_gradientn(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    colors = colors,
    name = "JI"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(angle = 90, size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_blank()
  ) +
  labs(x = "", y = "") +
  ggtitle("Receptors")

panel_c

# save panel
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_receptors_inhibitory.png"
  ),
  width = 150,
  height = 150,
  units = "mm",
  res = 300
)
panel_c
dev.off()
```

# Panel S3D
```{r}
# prepare as before but for targets
jaccard_targets_df <- data.frame(
  cell_types = names(jaccard_results_targets),
  jaccard_index = unlist(jaccard_results_targets),
  stringsAsFactors = FALSE
)

jaccard_targets_df <- jaccard_targets_df %>%
  separate(cell_types,
    into = c(
      "receiver",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

jaccard_targets_df <- jaccard_targets_df %>%
  spread(
    key = celltype2,
    value = jaccard_index
  )

# prep inhibitory neuron input
jaccard_targets_in <- jaccard_targets_df %>%
  filter(receiver == "Inhibitory Neurons")

jaccard_in_mat <- as.matrix(jaccard_targets_in[-c(1:2)])
rownames(jaccard_in_mat) <- jaccard_targets_in[[2]]

# plot panel D
panel_d <- ggcorrplot(jaccard_in_mat,
  outline.color = "white",
  lab = TRUE,
  lab_size = 5,
  type = "lower"
) +
  scale_fill_gradientn(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    colors = colors,
    name = "JI"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(angle = 90, size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 14),
    panel.grid.major = element_blank()
  ) +
  labs(x = "", y = "") +
  ggtitle("Targets")


panel_d

# save panel d
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_targets_inhibitory.png"
  ),
  width = 150,
  height = 150,
  units = "mm",
  res = 300
)
panel_d
dev.off()
```

# Panel S3E
```{r}
# prep excitatory neuron input for ligands
jaccard_ligands_ex <- jaccard_ligands_df %>%
  filter(receiver == "Excitatory Neurons")

jaccard_ex_mat <- as.matrix(jaccard_ligands_ex[-c(1:2)])
rownames(jaccard_ex_mat) <- jaccard_ligands_ex[[2]]

# plot
panel_e <- ggcorrplot(jaccard_ex_mat,
  outline.color = "white",
  lab = TRUE,
  lab_size = 5,
  type = "lower"
) +
  scale_fill_gradientn(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    colors = colors,
    name = "JI"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(angle = 90, size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_blank()
  ) +
  labs(y = "Excitatory Neurons", x = "") +
  ggtitle("Ligands")

panel_e

# save figure
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_ligands_excitatory.png"
  ),
  width = 150,
  height = 150,
  units = "mm",
  res = 300
)
panel_e
dev.off()
```

# Panel S3F
```{r}
# prep excitatory neuron input for receptors
jaccard_receptors_ex <- jaccard_receptors_df %>%
  filter(receiver == "Excitatory Neurons")

jaccard_ex_mat <- as.matrix(jaccard_receptors_ex[-c(1:2)])
rownames(jaccard_ex_mat) <- jaccard_receptors_ex[[2]]

# plot panel f
panel_f <- ggcorrplot(jaccard_ex_mat,
  outline.color = "white",
  lab = TRUE,
  lab_size = 5,
  type = "lower"
) +
  scale_fill_gradientn(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    colors = colors,
    name = "JI"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(angle = 90, size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_blank()
  ) +
  labs(x = "", y = "") +
  ggtitle("Receptors")

panel_f

# save panel
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_receptors_excitatory.png"
  ),
  width = 150,
  height = 150,
  units = "mm",
  res = 300
)
panel_f
dev.off()
```

# Panel S3G
```{r}
# prep excitatory neuron input for targets
jaccard_targets_ex <- jaccard_targets_df %>%
  filter(receiver == "Excitatory Neurons")

jaccard_ex_mat <- as.matrix(jaccard_targets_ex[-c(1:2)])
rownames(jaccard_ex_mat) <- jaccard_targets_ex[[2]]

# plot
panel_g <- ggcorrplot(jaccard_ex_mat,
  outline.color = "white",
  lab = TRUE,
  lab_size = 5,
  type = "lower"
) +
  scale_fill_gradientn(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    colors = colors,
    name = "JI"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(angle = 90, size = 14, face = "bold"),
    axis.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 14),
    panel.grid.major = element_blank()
  ) +
  labs(x = "", y = "") +
  ggtitle("Targets")

panel_g

# save panel G
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_targets_excitatory.png"
  ),
  width = 150,
  height = 150,
  units = "mm",
  res = 300
)
panel_g
dev.off()
```

# Compile Figure
* Bottom 6 panels are arranged first before being combined with the lollipop plot (panel A)
```{r}
bottom <- ggarrange(
  panel_b + theme(
    axis.title.y = element_textbox(
      linetype = 1,
      box.color = "darkslategray3",
      color = "black",
      fill = alpha("darkslategray3", alpha = 0.5),
      orientation = "left-rotated",
      r = grid::unit(3, "pt"),
      padding = margin(3, 5, 2, 5)
    ),
    plot.margin = unit(c(0, 0.5, 0, 0.5), "cm")
  ),
  panel_c + theme(
    plot.margin = unit(c(0, 0.5, 0, 0), "cm")
  ),
  panel_d + theme(
    plot.margin = unit(c(0, 0.5, 0, 0), "cm")
  ),
  panel_e + theme(
    plot.margin = unit(c(1, 0.5, 0, 0), "cm"),
    axis.title.y = element_textbox(
      linetype = 1,
      box.color = "slateblue3",
      color = "black",
      fill = alpha("slateblue3", alpha = 0.5),
      orientation = "left-rotated",
      r = grid::unit(3, "pt"),
      padding = margin(3, 5, 2, 5),
      margin = margin(0, 0, 5, 0)
    )
  ),
  panel_f + theme(
    plot.margin = unit(c(1, 0.5, 0, 0), "cm")
  ),
  panel_g + theme(
    plot.margin = unit(c(1, 0.5, 0, 0), "cm")
  ),
  ncol = 3,
  nrow = 2,
  common.legend = TRUE,
  legend = "right",
  labels = c("B", "C", "D", "E", "F", "G"),
  align = "hv"
)

# compile the whole figure and plot
fig_s3 <- plot_grid(lollipop,
  bottom,
  ncol = 1,
  rel_heights = c(1:2),
  labels = "A"
)

fig_s3
```

# Save compiled figure
```{r}
png(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "supp_figure3.png"
  ),
  width = 400,
  height = 400,
  units = "mm",
  res = 300
)
fig_s3
dev.off()
```

# Save compiled corr plots
```{r}
# save figure
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "corr_JI_all.png"
  ),
  width = 450,
  height = 300,
  units = "mm",
  res = 300
)
bottom
dev.off()
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2          styler_1.9.1         here_1.0.1           ggtext_0.1.2        
 [5] ggpubr_0.6.0         ggcorrplot_0.1.4     cowplot_1.1.1        patchwork_1.1.2     
 [9] multinichenetr_1.0.3 lubridate_1.9.2      forcats_1.0.0        stringr_1.5.0       
[13] dplyr_1.1.1          purrr_1.0.1          readr_2.1.4          tidyr_1.3.0         
[17] tibble_3.2.1         ggplot2_3.4.2        tidyverse_2.0.0     

loaded via a namespace (and not attached):
  [1] rsvd_1.0.5                  Hmisc_5.0-1                 ica_1.0-3                  
  [4] ps_1.7.5                    class_7.3-21                rprojroot_2.0.3            
  [7] foreach_1.5.2               lmtest_0.9-40               crayon_1.5.2               
 [10] rbibutils_2.2.13            MASS_7.3-58.3               nlme_3.1-162               
 [13] backports_1.4.1             sva_3.46.0                  rlang_1.1.0                
 [16] XVector_0.38.0              caret_6.0-94                ROCR_1.0-11                
 [19] irlba_2.3.5.1               callr_3.7.3                 nloptr_2.0.3               
 [22] limma_3.54.2                scater_1.26.1               BiocParallel_1.32.6        
 [25] rjson_0.2.21                bit64_4.0.5                 glue_1.6.2                 
 [28] sctransform_0.4.1           processx_3.8.1              pbkrtest_0.5.2             
 [31] parallel_4.2.3              vipor_0.4.5                 spatstat.sparse_3.0-1      
 [34] AnnotationDbi_1.60.2        BiocGenerics_0.44.0         UpSetR_1.4.0               
 [37] muscat_1.12.1               spatstat.geom_3.1-0         tidyselect_1.2.0           
 [40] SummarizedExperiment_1.28.0 SeuratObject_4.1.3          fitdistrplus_1.1-8         
 [43] variancePartition_1.28.9    XML_3.99-0.14               zoo_1.8-12                 
 [46] xtable_1.8-4                magrittr_2.0.3              evaluate_0.20              
 [49] Rdpack_2.4                  scuttle_1.8.4               cli_3.6.1                  
 [52] zlibbioc_1.44.0             rstudioapi_0.14             miniUI_0.1.1.1             
 [55] sp_1.6-0                    rpart_4.1.19                aod_1.3.2                  
 [58] locfdr_1.1-8                shiny_1.7.4                 BiocSingular_1.14.0        
 [61] xfun_0.38                   clue_0.3-64                 cluster_2.1.4              
 [64] caTools_1.18.2              tidygraph_1.2.3             KEGGREST_1.38.0            
 [67] clusterGeneration_1.3.7     ggrepel_0.9.3               listenv_0.9.0              
 [70] Biostrings_2.66.0           png_0.1-8                   future_1.32.0              
 [73] ipred_0.9-14                withr_2.5.0                 bitops_1.0-7               
 [76] ggforce_0.4.1               plyr_1.8.8                  hardhat_1.3.0              
 [79] dqrng_0.3.0                 e1071_1.7-13                pROC_1.18.0                
 [82] coda_0.19-4                 pillar_1.9.0                nichenetr_2.0.4            
 [85] gplots_3.1.3                GlobalOptions_0.1.2         cachem_1.0.7               
 [88] GetoptLong_1.0.5            RUnit_0.4.32                DelayedMatrixStats_1.20.0  
 [91] vctrs_0.6.2                 ellipsis_0.3.2              generics_0.1.3             
 [94] lava_1.7.2.1                tools_4.2.3                 foreign_0.8-84             
 [97] remaCor_0.0.11              beeswarm_0.4.0              munsell_0.5.0              
[100] tweenr_2.0.2                emmeans_1.8.5               proxy_0.4-27               
[103] DelayedArray_0.24.0         fastmap_1.1.1               compiler_4.2.3             
[106] abind_1.4-5                 httpuv_1.6.9                plotly_4.10.1              
[109] GenomeInfoDbData_1.2.9      prodlim_2023.03.31          gridExtra_2.3              
[112] glmmTMB_1.1.7               edgeR_3.40.2                lattice_0.21-8             
[115] ggnewscale_0.4.8            deldir_1.0-6                visNetwork_2.1.2           
[118] utf8_1.2.3                  later_1.3.0                 recipes_1.0.5              
[121] jsonlite_1.8.4              scales_1.2.1                ScaledMatrix_1.6.0         
[124] carData_3.0-5               pbapply_1.7-0               sparseMatrixStats_1.10.0   
[127] estimability_1.4.1          genefilter_1.80.3           lazyeval_0.2.2             
[130] promises_1.2.0.1            car_3.1-2                   doParallel_1.0.17          
[133] R.utils_2.12.2              goftest_1.2-3               spatstat.utils_3.0-2       
[136] reticulate_1.28             checkmate_2.1.0             cyclocomp_1.1.0            
[139] rmarkdown_2.21              blme_1.0-5                  statmod_1.5.0              
[142] Rtsne_0.16                  Biobase_2.58.0              uwot_0.1.14                
[145] igraph_1.4.2                survival_3.5-5              numDeriv_2016.8-1.1        
[148] yaml_2.3.7                  htmltools_0.5.5             memoise_2.0.1              
[151] Seurat_4.3.0                locfit_1.5-9.7              graphlayouts_0.8.4         
[154] IRanges_2.32.0              viridisLite_0.4.1           digest_0.6.31              
[157] RhpcBLASctl_0.23-42         commonmark_1.9.0            mime_0.12                  
[160] RSQLite_2.3.1               future.apply_1.10.0         remotes_2.4.2              
[163] data.table_1.14.8           blob_1.2.4                  R.oo_1.25.0                
[166] S4Vectors_0.36.2            DiagrammeR_1.0.9            labeling_0.4.2             
[169] splines_4.2.3               Formula_1.2-5               gridtext_0.1.5             
[172] RCurl_1.98-1.12             broom_1.0.4                 hms_1.1.3                  
[175] colorspace_2.1-0            base64enc_0.1-3             ggbeeswarm_0.7.1           
[178] GenomicRanges_1.50.2        shape_1.4.6                 nnet_7.3-18                
[181] Rcpp_1.0.10                 RANN_2.6.1                  mvtnorm_1.1-3              
[184] circlize_0.4.15             fansi_1.0.4                 tzdb_0.3.0                 
[187] parallelly_1.35.0           ModelMetrics_1.2.2.2        R6_2.5.1                   
[190] factoextra_1.0.7            ggridges_0.5.4              lifecycle_1.0.3            
[193] bluster_1.8.0               ggsignif_0.6.4              minqa_1.2.5                
[196] leiden_0.4.3                Matrix_1.5-4                desc_1.4.2                 
[199] RcppAnnoy_0.0.20            RColorBrewer_1.1-3          iterators_1.0.14           
[202] spatstat.explore_3.1-0      TMB_1.9.4                   gower_1.0.1                
[205] R.cache_0.16.0              htmlwidgets_1.6.2           markdown_1.6               
[208] beachmat_2.14.2             polyclip_1.10-4             shadowtext_0.1.2           
[211] timechange_0.2.0            mgcv_1.8-42                 ComplexHeatmap_2.14.0      
[214] globals_0.16.2              htmlTable_2.4.1             spatstat.random_3.1-4      
[217] progressr_0.13.0            codetools_0.2-19            matrixStats_0.63.0         
[220] metapod_1.6.0               randomForest_4.7-1.1        gtools_3.9.4               
[223] prettyunits_1.1.1           SingleCellExperiment_1.20.1 R.methodsS3_1.8.2          
[226] GenomeInfoDb_1.34.9         gtable_0.3.3                DBI_1.1.3                  
[229] stats4_4.2.3                tensor_1.5                  httr_1.4.5                 
[232] KernSmooth_2.23-20          stringi_1.7.12              progress_1.2.2             
[235] reshape2_1.4.4              farver_2.1.1                annotate_1.76.0            
[238] viridis_0.6.2               rex_1.2.1                   fdrtool_1.2.17             
[241] timeDate_4022.108           xml2_1.3.3                  boot_1.3-28.1              
[244] BiocNeighbors_1.16.0        lme4_1.1-32                 geneplotter_1.76.0         
[247] scattermore_0.8             scran_1.26.2                DESeq2_1.38.3              
[250] bit_4.0.5                   MatrixGenerics_1.10.0       spatstat.data_3.0-1        
[253] ggraph_2.1.0                pkgconfig_2.0.3             lmerTest_3.1-3             
[256] rstatix_0.7.2               knitr_1.42

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_S3.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_S3.Rmd"
))
```
