---
title: "figure_3"
author: "Tabea M. Soelter"
date: "2024-01-15"
output: html_document
---
**Plotting Figure 3**

__Goal__: This is a script for generating figure 3 of my manuscript.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
  * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*MultiNicheNet object*
* Name: multinichenet_output.rds
* Location: data/ccc/

__Analysis Plan__:
* Load necessary packages
* Load data
  * MultiNicheNet object
* Plot individual panels:
  * A: Grouped bar plot for the number of interactions by condition and time point
  * B: Upset plot to show overlap and unique interactions with stacked bar plot for receivers
  * C: Stacked bar plot of the interactions shared across time points and conditions
  * D: Stacked bar plot of interactions unique to the 6m time point
  * E: Stacked bar plot of interactions unique to the 12m time point
* Compile figure
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(ComplexUpset)
  library(ggpattern)
  library(patchwork)
  library(cowplot)
  library(ggpubr)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
output <- readRDS(here("data", "ccc", "multinichenet_output.rds"))
```

# Colors
```{r}
colors <- c(
  `Oligodendrocytes` = "royalblue3",
  `Microglia` = "orchid3",
  `OPCs` = "mediumpurple1",
  `Astrocytes` = "cornflowerblue",
  `Excitatory Neurons` = "slateblue3",
  `Inhibitory Neurons` = "darkslategray3",
  `Pericytes` = "deepskyblue3",
  `Endothelial cells` = "darkolivegreen4",
  `Meningeal cells` = "seagreen4",
  `Ependymal cells` = "lightcoral",
  `Fibroblasts` = "olivedrab3",
  `RGCs` = "violetred2"
)
```

# Set filtering criteria
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# vector of groups
group_oi <- c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
```

# Filter MultiNicheNet output
```{r}
output_filt <- filter_nichenet(output)
```

# Figure 3A
* Group bar chart of the number of interactions across senders by group (condition/time point)
```{r}
# Summarize interactions by sender and group; create condition column
prior_num <- output_filt %>%
  group_by(sender, group) %>%
  summarize(interactions = n_distinct(id_target)) %>%
  mutate(condition = ifelse(group %in% c("X6mAD", "X12mAD"), "AD", "WT"))

# Create time point column
prior_num$time_point <- sub("X(\\d+m)(AD|WT)", "\\1", prior_num$group)

# Factorize time_point column to ensure 6m comes before 12m
prior_num$time_point <- factor(prior_num$time_point, levels = c("6m", "12m"))

# Factorize condition column to ensure WT comes before AD
prior_num$condition <- factor(prior_num$condition, levels = c("WT", "AD"))

# plot grouped bar plot of interactions
grouped_bar <- ggplot(
  data = prior_num,
  aes(
    x = sender,
    y = interactions,
    fill = condition,
    pattern = time_point
  )
) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(preserve = "single"),
    color = "black",
    pattern_fill = "black",
    pattern_angle = 45,
    pattern_density = 0.1,
    pattern_spacing = 0.025,
    pattern_key_scale_factor = 0.6
  ) +
  geom_text(aes(label = interactions, interactions = interactions + 0.05),
    position = position_dodge(0.9),
    vjust = -0.3,
    fontface = "bold",
    size = 3.5
  ) +
  expand_limits(y = 650) +
  scale_fill_manual(values = c(
    `AD` = "#A6CEE3",
    `WT` = "#1F78B4"
  )) +
  scale_pattern_manual(values = c(
    `6m` = "stripe",
    `12m` = "none"
  )) +
  labs(
    x = "Sender",
    y = "# of Interactions",
    pattern = "Time Point",
    fill = "Condition"
  ) +
  guides(
    pattern = guide_legend(override.aes = list(fill = "white")),
    fill = guide_legend(override.aes = list(pattern = "none"))
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(face = "bold", color = "black", size = 12),
    axis.text.x = element_text(face = "bold", color = "black", size = 12),
    axis.title = element_text(face = "bold", color = "black", size = 14),
    legend.text = element_text(face = "bold", color = "black", size = 10),
    legend.title = element_text(face = "bold", color = "black", size = 12)
  )

grouped_bar

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "grouped_barplot.png"
  ),
  width = 300,
  height = 150,
  units = "mm",
  res = 300
)
grouped_bar
dev.off()
```

# Figure 3B
* Upset plot showing unique and shared interactions including receiver cell type
```{r}
# grab necessary info from filtered MultiNicheNet object
upset_df <- output_filt %>%
  select(sender, receiver, id_target, group)

# get columns of interest
upset_df_filt <- table(upset_df$id_target, upset_df$group)

# prepare upset df
upset_df_filt <- as.data.frame(upset_df_filt) %>%
  pivot_wider(
    names_from = Var2,
    values_from = Freq
  ) %>%
  as.data.frame() %>%
  mutate(Receiver = str_extract(
    Var1,
    "(?<=_)([[:alnum:]]+\\.[[:alnum:]]+)(?=_)"
  )) %>%
  as.data.frame()

# rename receivers (remove period)
upset_df_filt$Receiver <- gsub(
  "Excitatory.Neurons",
  "Excitatory Neurons",
  upset_df_filt$Receiver
)
upset_df_filt$Receiver <- gsub(
  "Inhibitory.Neurons",
  "Inhibitory Neurons",
  upset_df_filt$Receiver
)

# rename group (i.e. X6mAD etc) columns (remove X)
colnames(upset_df_filt) <- sub("^X", "", colnames(upset_df_filt))

# pull out group columns (6mAD, 12mAD, 6mWT, 12mWT)
groups <- colnames(upset_df_filt)[2:5]

# changing 0/1  format to True/False for the group columns
upset_df_filt[groups] <- upset_df_filt[groups] == 1

# plot the upset plot with a stacked intersection bar plot
upset <- upset(upset_df_filt,
  groups,
  name = "Group",
  base_annotations = list(
    "Intersection Size" = intersection_size(
      counts = TRUE,
      mapping = aes(fill = Receiver),
      bar_number_threshold = 1,
      text = list(color = "black", fontface = "bold")
    ) +
      theme_bw() +
      theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_text(
          size = 14,
          face = "bold"
        ),
        legend.text = element_text(
          size = 12,
          face = "bold"
        ),
        axis.title.y = element_text(
          size = 14,
          face = "bold"
        )
      ) +
      expand_limits(y = 1700) +
      scale_fill_manual(values = colors)
  ),
  themes = upset_default_themes(
    text = element_text(
      face = "bold",
      color = "black"
    ),
    axis.text.y = element_text(
      color = "black",
      face = "bold",
      size = 12
    ),
    axis.title = element_text(size = 14)
  ),
  set_sizes = upset_set_size(position = "right") +
    geom_text(aes(label = ..count..),
      hjust = -0.1,
      stat = "count",
      size = 3.5,
      fontface = "bold"
    ) +
    expand_limits(y = 2100) +
    theme(
      axis.text.x = element_text(color = "black", size = 12),
      axis.title.x = element_text(size = 14),
      axis.text.y = element_blank()
    ),
  guides = "over"
) +
  theme(
    text = element_text(face = "bold", color = "black"),
    axis.title = element_text(size = 11)
  )

upset

png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "upset.png"
  ),
  width = 300,
  height = 150,
  units = "mm",
  res = 300
)
upset
dev.off()
```

# Figure 3C
* Stacked bar plot of shared interactions
```{r}
# grab/create necessary columns
all <- output_filt %>%
  select(id_target, sender, group) %>%
  mutate(condition = sub("^X", "", group)) %>%
  select(-group)

# filter for shared interactions
shared <- all %>%
  group_by(id_target) %>%
  filter(n() > 1)

# change condition info to just AD/WT
shared$condition <- substring(
  shared$condition,
  nchar(shared$condition) - 1,
  nchar(shared$condition)
)

# remove duplicates
shared <- unique(shared)

# prepare shared bar plot summary df
sums <- shared %>%
  group_by(sender, condition) %>%
  summarize(interactions = n_distinct(id_target))

# plot shared stacked bar plot
shared_bar <- ggplot(
  sums,
  aes(x = sender, y = interactions, fill = condition)
) +
  geom_bar(stat = "identity", position = "stack") +
  xlab("Sender") +
  ylab("# of Interactions") +
  scale_fill_discrete(name = "Condition") +
  theme_bw() +
  geom_text(aes(label = interactions),
    position = "stack",
    vjust = 1.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(`AD` = "#A6CEE3", `WT` = "#1F78B4")) +
  labs(fill = "Condition") +
  theme(
    legend.text = element_text(face = "bold"),
    axis.text = element_text(face = "bold", size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12),
    text = element_text(face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 12,
      color = "black"
    )
  )

shared_bar

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "barplot_shared_senders.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
shared_bar
dev.off()
```

# Figure 3D
* Stacked bar plot of 6m-specific interactions
```{r}
# grab LRTs shared in AD
shared_int_ad <- shared %>%
  filter(condition == "AD") %>%
  select(id_target) %>%
  unique()

shared_int_ad <- shared_int_ad$id_target

# grab LRTs shared in WT
shared_int_wt <- shared %>%
  filter(condition == "WT") %>%
  select(id_target) %>%
  unique()

shared_int_wt <- shared_int_wt$id_target

# filter out any interactions that are shared by condition
all_filt <- all %>%
  filter(!(condition %in% c("6mAD", "12mAD") & id_target %in% shared_int_ad) &
    !(condition %in% c("6mWT", "12mWT") & id_target %in% shared_int_wt))

# get 6m time point interactions for both conditions
filt_6m <- all_filt %>%
  filter(condition %in% c("6mAD", "6mWT"))

# aggregate numbers by sender and condition for plotting
prior_num <- filt_6m %>%
  group_by(sender, condition) %>%
  summarize(interactions = n_distinct(id_target))

# plot stacked 6m bar plot, split by condition
unique_6m_bar <- ggplot(
  prior_num,
  aes(x = sender, y = interactions, fill = condition)
) +
  geom_bar(stat = "identity", position = "stack") +
  xlab("Sender") +
  ylab("# of Interactions") +
  scale_fill_discrete(name = "Condition") +
  theme_bw() +
  geom_text(aes(label = interactions),
    position = "stack",
    vjust = 1.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(`6mAD` = "#A6CEE3", `6mWT` = "#1F78B4")) +
  labs(fill = "Condition") +
  theme(
    legend.text = element_text(face = "bold"),
    axis.text = element_text(face = "bold", size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12),
    text = element_text(face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 12,
      color = "black"
    )
  )

unique_6m_bar

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "barplot_unique_6m.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
unique_6m_bar
dev.off()
```

# Figure 3E
* Stacked bar plot of 12m-specific interactions
```{r}
# get 12m specific interactions
filt_12m <- all_filt %>%
  filter(condition %in% c("12mAD", "12mWT"))

# aggregate numbers by sender and condition for plotting
prior_num <- filt_12m %>%
  group_by(sender, condition) %>%
  summarize(interactions = n_distinct(id_target))

# plot stacked 12m bar plot, split by condition
unique_12m_bar <- ggplot(
  prior_num,
  aes(x = sender, y = interactions, fill = condition)
) +
  geom_bar(stat = "identity", position = "stack") +
  xlab("Sender") +
  ylab("# of Interactions") +
  scale_fill_discrete(name = "Condition") +
  theme_bw() +
  geom_text(aes(label = interactions),
    position = "stack",
    vjust = 1.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(`12mAD` = "#A6CEE3", `12mWT` = "#1F78B4")) +
  labs(fill = "Condition") +
  theme(
    legend.text = element_text(face = "bold"),
    axis.text = element_text(face = "bold", size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12),
    text = element_text(face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 12,
      color = "black"
    )
  )

unique_12m_bar

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "barplot_unique_12m.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
unique_12m_bar
dev.off()
```

# Compile Figure
* To get the right proportions, I broke the figure up into thirds (top, middle, bottom)
```{r}
top <- ggarrange(grouped_bar,
  labels = c("A")
)

middle <- ggarrange(upset,
  labels = c("B")
)

bottom <- ggarrange(shared_bar,
  unique_6m_bar,
  unique_12m_bar,
  ncol = 3,
  labels = c("C", "D", "E")
)

# compile whole figure
fig3 <- ggarrange(top,
  middle,
  bottom,
  nrow = 3,
  heights = c(1, 1.7, 1.3)
)

# save figure
png(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "figure3.png"
  ),
  width = 300,
  height = 325,
  units = "mm",
  res = 300
)
fig3
dev.off()
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2        styler_1.9.1       here_1.0.1         ggpubr_0.6.0       cowplot_1.1.1     
 [6] patchwork_1.1.2    ggpattern_1.1.0-0  ComplexUpset_1.3.5 SeuratObject_4.1.3 Seurat_4.3.0      
[11] lubridate_1.9.2    forcats_1.0.0      stringr_1.5.0      dplyr_1.1.1        purrr_1.0.1       
[16] readr_2.1.4        tidyr_1.3.0        tibble_3.2.1       ggplot2_3.4.2      tidyverse_2.0.0   

loaded via a namespace (and not attached):
  [1] backports_1.4.1        plyr_1.8.8             igraph_1.4.2           lazyeval_0.2.2        
  [5] sp_1.6-0               splines_4.2.3          listenv_0.9.0          scattermore_0.8       
  [9] digest_0.6.31          htmltools_0.5.5        fansi_1.0.4            memoise_2.0.1         
 [13] magrittr_2.0.3         tensor_1.5             cluster_2.1.4          ROCR_1.0-11           
 [17] tzdb_0.3.0             remotes_2.4.2          globals_0.16.2         matrixStats_0.63.0    
 [21] R.utils_2.12.2         timechange_0.2.0       spatstat.sparse_3.0-1  colorspace_2.1-0      
 [25] ggrepel_0.9.3          xfun_0.38              callr_3.7.3            crayon_1.5.2          
 [29] jsonlite_1.8.4         progressr_0.13.0       spatstat.data_3.0-1    survival_3.5-5        
 [33] zoo_1.8-12             glue_1.6.2             polyclip_1.10-4        gtable_0.3.3          
 [37] leiden_0.4.3           R.cache_0.16.0         car_3.1-2              future.apply_1.10.0   
 [41] abind_1.4-5            scales_1.2.1           DBI_1.1.3              rstatix_0.7.2         
 [45] spatstat.random_3.1-4  miniUI_0.1.1.1         Rcpp_1.0.10            viridisLite_0.4.1     
 [49] xtable_1.8-4           units_0.8-1            reticulate_1.28        proxy_0.4-27          
 [53] htmlwidgets_1.6.2      rex_1.2.1              httr_1.4.5             RColorBrewer_1.1-3    
 [57] ellipsis_0.3.2         ica_1.0-3              farver_2.1.1           pkgconfig_2.0.3       
 [61] R.methodsS3_1.8.2      uwot_0.1.14            deldir_1.0-6           utf8_1.2.3            
 [65] labeling_0.4.2         tidyselect_1.2.0       rlang_1.1.0            reshape2_1.4.4        
 [69] later_1.3.0            cachem_1.0.7           munsell_0.5.0          tools_4.2.3           
 [73] cli_3.6.1              generics_0.1.3         broom_1.0.4            ggridges_0.5.4        
 [77] evaluate_0.20          fastmap_1.1.1          yaml_2.3.7             goftest_1.2-3         
 [81] processx_3.8.1         knitr_1.42             fitdistrplus_1.1-8     RANN_2.6.1            
 [85] pbapply_1.7-0          future_1.32.0          nlme_3.1-162           mime_0.12             
 [89] R.oo_1.25.0            xml2_1.3.3             gridpattern_1.0.2      compiler_4.2.3        
 [93] rstudioapi_0.14        plotly_4.10.1          png_0.1-8              e1071_1.7-13          
 [97] ggsignif_0.6.4         spatstat.utils_3.0-2   stringi_1.7.12         cyclocomp_1.1.0       
[101] ps_1.7.5               desc_1.4.2             lattice_0.21-8         Matrix_1.5-4          
[105] classInt_0.4-9         vctrs_0.6.2            pillar_1.9.0           lifecycle_1.0.3       
[109] spatstat.geom_3.1-0    lmtest_0.9-40          RcppAnnoy_0.0.20       data.table_1.14.8     
[113] irlba_2.3.5.1          httpuv_1.6.9           R6_2.5.1               promises_1.2.0.1      
[117] KernSmooth_2.23-20     gridExtra_2.3          parallelly_1.35.0      codetools_0.2-19      
[121] MASS_7.3-58.3          rprojroot_2.0.3        withr_2.5.0            sctransform_0.4.1     
[125] parallel_4.2.3         hms_1.1.3              grid_4.2.3             class_7.3-21          
[129] rmarkdown_2.21         carData_3.0-5          Rtsne_0.16             sf_1.0-12             
[133] spatstat.explore_3.1-0 shiny_1.7.4

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_3.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_3.Rmd"
))
```
