---
title: "figure_4"
author: "Tabea M. Soelter"
date: "2024-01-21"
output: html_document
---
**Plotting Main Figure 4**

__Goal__: This is a script for generating figure 4 of my manuscript.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

#__Data__: EDIT THIS 
*MultiNicheNet object*
* Name: multinichenet_output.rds
* Location: data/ccc/

#__Analysis Plan__: EDIT THIS - maybe not condition for heatmap ... we will see
* Load necessary packages
* Load data
  * MultiNicheNet object
* Plot individual panels:
  * A: Heatmap of pseudobulk expression annotated by receiver, condition, and time point
  * B: Bubble plot for6-month FEA
  * C: Bubble plot for 12-month FEA
* Compile figure
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(ComplexHeatmap)
  library(RColorBrewer)
  library(circlize)
  library(patchwork)
  library(cowplot)
  library(ggpubr)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data

-   DESeq2 inputs

    -   inhibitory and excitatory neurons

```{r}
    excitatory_neurons_group_6mAD_vs_6mWT_sig_genes <- read_csv("results/intermediate_outputs/03_dea/excitatory_neurons_group_6mAD_vs_6mWT_sig_genes.csv")

    excitatory_neurons_group_12mAD_vs_12mWT_sig_genes <- read_csv("results/intermediate_outputs/03_dea/excitatory_neurons_group_12mAD_vs_12mWT_sig_genes.csv")

    # inhibitory neurons
    inhibitory_neurons_group_6mAD_vs_6mWT_sig_genes <- read_csv("results/intermediate_outputs/03_dea/inhibitory_neurons_group_6mAD_vs_6mWT_sig_genes.csv")

inhibitory_neurons_group_12mAD_vs_12mWT_sig_genes <- read_csv("results/intermediate_outputs/03_dea/inhibitory_neurons_group_12mAD_vs_12mWT_sig_genes.csv")
```

# Prepare dfs

```{r}
# add time point information
excitatory_neurons_group_6mAD_vs_6mWT_sig_genes$time_point <- "6m"
excitatory_neurons_group_12mAD_vs_12mWT_sig_genes$time_point <- "12m"
inhibitory_neurons_group_6mAD_vs_6mWT_sig_genes$time_point <- "6m"
inhibitory_neurons_group_12mAD_vs_12mWT_sig_genes$time_point <- "12m"

# add receiver information
excitatory_neurons_group_6mAD_vs_6mWT_sig_genes$receiver <- "excitatory_neurons"
excitatory_neurons_group_12mAD_vs_12mWT_sig_genes$receiver <- "excitatory_neurons"
inhibitory_neurons_group_6mAD_vs_6mWT_sig_genes$receiver <- "inhibitory_neurons"
inhibitory_neurons_group_12mAD_vs_12mWT_sig_genes$receiver <- "inhibitory_neurons"

# combine all 4 dfs into 1 for plotting purposes
combined_df <- rbind(excitatory_neurons_group_6mAD_vs_6mWT_sig_genes,
                     excitatory_neurons_group_12mAD_vs_12mWT_sig_genes,
                     inhibitory_neurons_group_6mAD_vs_6mWT_sig_genes,
                     inhibitory_neurons_group_12mAD_vs_12mWT_sig_genes) %>%
  select(gene, log2FoldChange, pvalue, time_point, receiver) %>%
  pivot_wider(
    id_cols = gene,
    names_from = c(time_point, receiver),
    values_from = c(log2FoldChange, pvalue)
  ) #%>%
  #mutate(gene_sig = case_when(
  #  pvalue_geo < 0.05 & pvalue_gse < 0.05 ~ paste0(gene, "**"),
  #  pvalue_geo < 0.05 ~ paste0(gene, "*"),
  #  pvalue_gse < 0.05 ~ paste0(gene, "^"),
  #  TRUE ~ gene
  #)) %>%
  #select(-c(pvalue_geo, pvalue_gse, gene)) %>%
  #column_to_rownames(var = "gene_sig")


# I am excluding the pvalue columns as all of these genes have a adjusted sig p value
combined_df <- rbind(excitatory_neurons_group_6mAD_vs_6mWT_sig_genes,
                     excitatory_neurons_group_12mAD_vs_12mWT_sig_genes,
                     inhibitory_neurons_group_6mAD_vs_6mWT_sig_genes,
                     inhibitory_neurons_group_12mAD_vs_12mWT_sig_genes) %>%
  select(gene, log2FoldChange, time_point, receiver) %>%
  pivot_wider(
    id_cols = gene,
    names_from = c(time_point, receiver),
    values_from = log2FoldChange
  ) %>%
  column_to_rownames(var = "gene")


# Prepare heatmap annotation
meta <- as.data.frame(colnames(combined_df))
rownames(meta) <- meta$`colnames(combined_df)`

receiver <- c("excitatory_neurons", "excitatory_neurons", "inhibitory_neurons", "inhibitory_neurons")
time_point <- c("6m", "12m", "6m", "12m")

meta$time_point <- time_point
meta$receiver <- receiver

meta <- meta[, 2:3]


anno_cols <- list(
  "receiver" = c(
    "excitatory_neurons" = "grey75",
    "inhibitory_neurons" = "grey36"
  ),
  "time_point" = c(
    "6m" = "lavender",
    "12m" = "purple"
  )
)

# Generate annotation
anno <- HeatmapAnnotation(
  df = meta,
  show_annotation_name = FALSE,
  col = anno_cols#,
  #annotation_legend_param = list(
  #  dataset = list(
  #    title = "Receiver",
  #    direction = "horizontal",
  #    #labels = c("Morabito", "Lau"),
  #    labels_gp = gpar(fontface = "bold")
  #  )
  )
#)

# Convert df to matrix
mat <- as.matrix(combined_df)

# Set heatmap colors
cols <- colorRamp2(c(-0.5, 0, 0.5), rev(brewer.pal(n = 3, name = "RdBu")))

# Plot heatmap
gex_heatmap <- Heatmap(mat,
  col = cols,
  top_annotation = anno,
  show_column_names = FALSE,
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  row_title = "Signaling Mediators",
  row_title_gp = gpar(fontface = "bold", fontsize = 12),
  #row_names_gp = gpar(col = row_colors, fontface = "bold"),
  heatmap_legend_param = list(
    direction = "horizontal",
    title = "Differential Expression",
    labels_gp = gpar(fontface = "bold")
  )
)
```






# Session info
```{r}
sessionInfo() # see output below
```

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_4.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_4.Rmd"
))
```
