---
title: "figure_S1"
author: "Tabea M. Soelter"
date: "2024-01-17"
output: html_document
---
**Plotting Supp Figure 1**

__Goal__: This is a script for generating supp figure 1 of my manuscript.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
  * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*MultiNicheNet object*
* Name: multinichenet_output.rds
* Location: data/ccc/

__Analysis Plan__:
* Load necessary packages
* Load data
  * MultiNicheNet object
* Plot individual panels:
  * A: Bar plot of interactions across all cell types
  * B: Bar plot of prioritized interactions w cell types of interest
  * C: Stacked bar plot of B colored by receiver cell types
* Compile figure S1
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
  library(cowplot)
  library(ggpubr)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
output <- readRDS(here("data", "ccc", "multinichenet_output.rds"))
```

# Colors
```{r}
colors <- c(
  `Oligodendrocytes` = "royalblue3",
  `Microglia` = "orchid3",
  `OPCs` = "mediumpurple1",
  `Astrocytes` = "cornflowerblue",
  `Excitatory Neurons` = "slateblue3",
  `Inhibitory Neurons` = "darkslategray3",
  `Pericytes` = "deepskyblue3",
  `Endothelial cells` = "darkolivegreen4",
  `Meningeal cells` = "seagreen4",
  `Ependymal cells` = "lightcoral",
  `Fibroblasts` = "olivedrab3",
  `RGCs` = "violetred2"
)
```

# Set Filtering Criteria
* These are the same as previously
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")
```

# Figure S1A
* Bar plot of all the interactions
```{r}
# filter MultiNicheNet output
total <- output$lr_target_prior_cor %>%
  inner_join(output$ligand_activities_targets_DEgenes$ligand_activities %>%
    distinct(ligand, target, direction_regulation, contrast)) %>%
  inner_join(contrast_tbl)

# filter for needed columns
total_filt <- total %>%
  select("id_target", "sender")

# compile number of interactions by sender
total_num <- total_filt %>%
  group_by(sender) %>%
  summarize(interactions = n_distinct(id_target)) %>%
  mutate(type = print("total"))

# replace names to remove periods
total_num$sender <- ifelse(
  total_num$sender == "Inhibitory.Neurons",
  "Inhibitory Neurons",
  ifelse(total_num$sender == "Excitatory.Neurons",
    "Excitatory Neurons",
    ifelse(total_num$sender == "Endothelial.cells",
      "Endothelial cells",
      ifelse(total_num$sender == "Meningeal.cells",
        "Meningeal cells",
        ifelse(total_num$sender == "Ependymal.cells",
          "Ependymal cells", total_num$sender
        )
      )
    )
  )
)

# plot bar plot of all interactions
panel_a <- ggplot(total_num, aes(x = sender, y = interactions, fill = sender)) +
  geom_bar(stat = "identity") +
  xlab("Sender") +
  ylab("# of all Interactions") +
  theme_bw() +
  geom_text(aes(label = interactions),
    position = "stack",
    vjust = 2,
    fontface = "bold"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.text = element_text(face = "bold", size = 10),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12, color = "black")
  ) +
  scale_fill_manual(values = colors, name = "Sender")

panel_a

# save panel
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "barplot_all.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
panel_a
dev.off()
```

# Figure S1B
* Bar plot of prioritized interactions between senders and receivers of interest
```{r}
# filter MultiNicheNet output and select columns needed for plotting
oi <- output$lr_target_prior_cor %>%
  inner_join(output$ligand_activities_targets_DEgenes$ligand_activities %>%
    distinct(ligand, target, direction_regulation, contrast)) %>%
  inner_join(contrast_tbl) %>%
  filter(receiver %in% receiver_oi, sender %in% sender_oi)

oi_filt <- oi %>%
  select("id_target", "sender")

# aggregate interactions by sender
oi_num <- oi_filt %>%
  group_by(sender) %>%
  summarize(interactions = n_distinct(id_target))

# plot bar plot of interactions with specified sender and receiver cell types
panel_b <- ggplot(oi_num, aes(x = sender, y = interactions, fill = sender)) +
  geom_bar(stat = "identity") +
  xlab("Sender") +
  ylab("# of Unique Interactions") +
  theme_bw() +
  geom_text(aes(label = interactions),
    position = "stack",
    vjust = 2,
    fontface = "bold"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.text = element_text(face = "bold", size = 10),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12, color = "black")
  ) +
  scale_fill_manual(values = colors, name = "Sender")

panel_b

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "barplot_oi.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
panel_b
dev.off()
```

# Figure S1C
* Stacked barplot of predicted interactions colored by receiver
```{r}
oi_filt_rec <- oi %>%
  select("id_target", "sender", "receiver")

oi_filt_rec$receiver <- ifelse(oi_filt_rec$receiver == "Excitatory.Neurons",
  "Excitatory Neurons",
  "Inhibitory Neurons"
)

# aggregate interactions by sender and receiver
oi_num_rec <- oi_filt_rec %>%
  group_by(sender, receiver) %>%
  summarize(interactions = n_distinct(id_target))

# plot stacked bar plot
panel_c <- ggplot(
  oi_num_rec,
  aes(x = sender, y = interactions, fill = receiver)
) +
  geom_bar(stat = "identity") +
  xlab("Sender") +
  ylab("# of Interactions") +
  theme_bw() +
  geom_text(aes(label = interactions),
    position = "stack",
    vjust = 2,
    fontface = "bold"
  ) +
  theme(
    legend.text = element_text(face = "bold", size = 10),
    axis.text = element_text(face = "bold", size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12)
  ) +
  scale_fill_manual(values = colors, name = "Receiver")

panel_c

# save plot
png(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "barplot_oi_by_receiver.png"
  ),
  width = 200,
  height = 175,
  units = "mm",
  res = 300
)
panel_c
dev.off()
```

# Compile Figure
```{r}
top <- ggarrange(panel_a,
  labels = "A"
)

bottom <- ggarrange(panel_b,
  panel_c,
  ncol = 2,
  labels = c("B", "C"),
  align = "h",
  widths = c(1, 1.5)
)

fig_s1 <- ggarrange(top,
  bottom,
  nrow = 2
)

fig_s1

# save plot
png(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "supp_figure1.png"
  ),
  width = 325,
  height = 200,
  units = "mm",
  res = 300
)
fig_s1
dev.off()
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] ggpubr_0.6.0    lintr_3.0.2     styler_1.9.1    here_1.0.1      cowplot_1.1.1  
 [6] patchwork_1.1.2 lubridate_1.9.2 forcats_1.0.0   stringr_1.5.0   dplyr_1.1.1    
[11] purrr_1.0.1     readr_2.1.4     tidyr_1.3.0     tibble_3.2.1    ggplot2_3.4.2  
[16] tidyverse_2.0.0

loaded via a namespace (and not attached):
 [1] ps_1.7.5          rprojroot_2.0.3   digest_0.6.31     utf8_1.2.3       
 [5] R6_2.5.1          backports_1.4.1   evaluate_0.20     pillar_1.9.0     
 [9] rlang_1.1.0       lazyeval_0.2.2    rstudioapi_0.14   car_3.1-2        
[13] callr_3.7.3       R.utils_2.12.2    R.oo_1.25.0       rmarkdown_2.21   
[17] desc_1.4.2        labeling_0.4.2    munsell_0.5.0     broom_1.0.4      
[21] compiler_4.2.3    xfun_0.38         pkgconfig_2.0.3   htmltools_0.5.5  
[25] tidyselect_1.2.0  fansi_1.0.4       crayon_1.5.2      tzdb_0.3.0       
[29] withr_2.5.0       R.methodsS3_1.8.2 grid_4.2.3        gtable_0.3.3     
[33] lifecycle_1.0.3   magrittr_2.0.3    scales_1.2.1      cli_3.6.1        
[37] stringi_1.7.12    carData_3.0-5     farver_2.1.1      ggsignif_0.6.4   
[41] remotes_2.4.2     rex_1.2.1         xml2_1.3.3        generics_0.1.3   
[45] vctrs_0.6.2       cyclocomp_1.1.0   tools_4.2.3       R.cache_0.16.0   
[49] glue_1.6.2        hms_1.1.3         yaml_2.3.7        processx_3.8.1   
[53] abind_1.4-5       fastmap_1.1.1     timechange_0.2.0  colorspace_2.1-0 
[57] rstatix_0.7.2     knitr_1.42

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: < 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_S1.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_S1.Rmd"
))
```
