---
title: "figure_S5"
author: "Tabea M. Soelter"
date: "2024-02-01"
output: html_document
---
**Plotting Supplementary Manuscript Figure 4**

__Goal__: This is a script for generating supp figure 4 of my manuscript, which includes an alluvial plot of the LRTs that have DE AD gene as a target.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__: 
*MultiNicheNet object*
* Name: multinichenet_output.rds
* Location: data/ccc/
*Filtered DESeq2 output*
* Name: filtered_dea_targets.rds
* Location: results/final_outputs/04_dea/
*Converted AD gene list*
* Name: ad_genes_mm.rds
* Location: doc/

__Analysis Plan__:
* Load necessary packages
* Load data
  * Load AD gene lists
  * MultiNicheNet object
  * Filtered DESeq2 output
* Plot figure S5:
  * Alluvial plot of LRTs that have a sig DE AD target gene
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggalluvial)
  library(patchwork)
  library(cowplot)
  library(ggpubr)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Color palette
```{r}
colors <- c(
  `Oligodendrocytes` = "royalblue3",
  `Microglia` = "orchid3",
  `OPCs` = "mediumpurple1",
  `Astrocytes` = "cornflowerblue",
  `Excitatory.Neurons` = "slateblue3",
  `Inhibitory.Neurons` = "darkslategray3",
  `Pericytes` = "deepskyblue3",
  `Endothelial cells` = "darkolivegreen4",
  `Meningeal cells` = "seagreen4",
  `Ependymal cells` = "lightcoral",
  `Fibroblasts` = "olivedrab3",
  `RGCs` = "violetred2"
)
```

# Load Data
* MultiNicheNet output
* AD gene list
* Filtered DEA df
```{r}
# MultiNicheNet
output <- readRDS(here("data", "ccc", "multinichenet_output.rds"))

# Mouse AD gene list
ad_genes_mm <- readRDS(here("doc", "ad_genes_mm.rds"))

# Target gene expression info from DESeq2
all_gex <- readRDS(here(
  "results",
  "final_outputs",
  "04_dea",
  "filtered_dea_targets.rds"
))
```

# Filter MultiNicheNet output
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# vector of groups of interest
group_oi <- c("X6mAD", "X6mWT", "X12mAD", "X12mWT")

# Filter output using above criteria and add time point column
output_filt <- filter_nichenet(output)
```

# Filter DEA info by AD-associated genes
```{r}
targets_ad <- all_gex %>%
  filter(gene %in% ad_genes_mm)

# filter by significance
sig_ad_targets <- targets_ad %>%
  filter(padj < 0.05)
```

# Prepare df for plotting
```{r}
# Filter MultiNicheNet object by AD genes
ad_lrt <- output_filt %>%
  select(sender, receiver, ligand, receptor, target, group) %>%
  filter(target %in% sig_ad_targets$gene) %>%
  unique()

# Factorize columns of interest (for plotting purposes)
vec <- c("ligand", "receptor", "target")

ad_lrt[vec] <- lapply(ad_lrt[vec], function(x) factor(x, levels = unique(x)))

# Reorder columns of interest by the sender to aid with legibility
ad_lrt$ligand <- reorder(
  ad_lrt$ligand,
  desc(ad_lrt$sender)
)

ad_lrt$receptor <- reorder(
  ad_lrt$receptor,
  desc(ad_lrt$sender)
)

ad_lrt$target <- reorder(
  ad_lrt$target,
  desc(ad_lrt$sender)
)
```

# Plot figure S5
```{r}
lrt_alluvial <- ggplot(ad_lrt,
  aes(axis1 = ligand, axis2 = receptor, axis3 = target),
  label = stratum
) +
  xlab("Type") +
  geom_alluvium() +
  stat_alluvium(lode.guidance = "frontback") +
  geom_stratum(aes(fill = sender)) +
  geom_label(
    stat = "stratum",
    fill = "white",
    aes(label = after_stat(stratum)),
    size = 3,
    fontface = "bold"
  ) +
  geom_flow(aes(fill = sender), stat = "alluvium", color = "black") +
  scale_fill_manual(values = colors) +
  scale_x_discrete(limits = c("Ligand", "Receptor", "Target")) +
  theme_void() +
  theme(
    axis.text.x = element_text(vjust = 2, face = "bold", size = 12),
    legend.position = "bottom",
    plot.margin = margin(0, -2, 0, -2, "cm"),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 14)
  ) +
  guides(fill = guide_legend(title = "Sender"))

lrt_alluvial
```

# Save figure S5
```{r}
ggsave(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "supp_figure5.png"
  ),
  lrt_alluvial,
  dpi = 300,
  width = 10,
  height = 10,
  bg = "#ffffff"
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 [9] LC_ADDRESS=C               LC_TELEPHONE=C            
[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2       styler_1.9.1      here_1.0.1        ggpubr_0.6.0     
 [5] cowplot_1.1.1     patchwork_1.1.2   ggalluvial_0.12.5 lubridate_1.9.2  
 [9] forcats_1.0.0     stringr_1.5.0     dplyr_1.1.1       purrr_1.0.1      
[13] readr_2.1.4       tidyr_1.3.0       tibble_3.2.1      ggplot2_3.4.2    
[17] tidyverse_2.0.0  

loaded via a namespace (and not attached):
 [1] carData_3.0-5     R.utils_2.12.2    diffobj_0.3.5     yaml_2.3.7       
 [5] remotes_2.4.2     pillar_1.9.0      backports_1.4.1   glue_1.6.2       
 [9] digest_0.6.31     ggsignif_0.6.4    colorspace_2.1-0  htmltools_0.5.5  
[13] R.oo_1.25.0       WriteXLS_6.4.0    pkgconfig_2.0.3   broom_1.0.4      
[17] scales_1.2.1      processx_3.8.1    tzdb_0.3.0        timechange_0.2.0 
[21] generics_0.1.3    farver_2.1.1      car_3.1-2         withr_2.5.0      
[25] lazyeval_0.2.2    cli_3.6.1         magrittr_2.0.3    crayon_1.5.2     
[29] evaluate_0.20     ps_1.7.5          R.methodsS3_1.8.2 fansi_1.0.4      
[33] R.cache_0.16.0    rstatix_0.7.2     xml2_1.3.3        textshaping_0.3.6
[37] tools_4.2.3       hms_1.1.3         cyclocomp_1.1.0   lifecycle_1.0.3  
[41] munsell_0.5.0     callr_3.7.3       rex_1.2.1         compiler_4.2.3   
[45] systemfonts_1.0.4 rlang_1.1.0       grid_4.2.3        rstudioapi_0.14  
[49] labeling_0.4.2    rmarkdown_2.21    waldo_0.4.0       gtable_0.3.3     
[53] abind_1.4-5       rematch2_2.1.2    R6_2.5.1          knitr_1.42       
[57] fastmap_1.1.1     utf8_1.2.3        rprojroot_2.0.3   ragg_1.2.5       
[61] desc_1.4.2        stringi_1.7.12    vctrs_0.6.2       tidyselect_1.2.0 
[65] xfun_0.38

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_S5.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_S5.Rmd"
))
```
