---
title: "figure_5"
author: "Tabea M. Soelter"
date: "2024-02-09"
output: html_document
---
**Plotting Manuscript Figure 5**

__Goal__: This is a script for generating figure 5 of my manuscript. We are also saving supp table 4.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*MultiNicheNet object*
* Name: multinichenet_output.rds
* Location: data/ccc/
*Filtered DESeq2 output*
* Name: filtered_dea_targets.rds
* Location: results/final_outputs/04_dea/
*AD Gene List*
* Name: ad_genes_mm.rds
* Location: doc/
*Network properties*
* Names: signaling_grn_properties_6m.rds, signaling_grn_properties_12m.rds
* Location: results/intermediate_outputs/05_grn/
*Unfiltered DESeq2 outputs*
* Names:
  1. excitatory_neurons_group_6mAD_vs_6mWT_all_genes.csv
  2. excitatory_neurons_group_12mAD_vs_12mWT_all_genes.csv
  3. inhibitory_neurons_group_6mAD_vs_6mWT_all_genes.csv
  4. inhibitory_neurons_group_12mAD_vs_12mWT_all_genes.csv
* Location: results/intermediate_outputs/03_dea/

#__Analysis Plan__: EDIT THIS
* Load necessary packages
* Load data
  * Filtered DESeq2 output
  * MultiNicheNet object
  * AD gene list
  * Network properties
  * DESeq2 outputs
* Filter all input data
* Plot individual panels:
  * A: Alluvial plot of LRs where targets are AD genes
  * B: Split Heatmap
      * Top: differential gex of AD-associated targets
      * Bottom: differential gex of mediators of AD-associated interactions
* Compile figure
* Save figure

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggalluvial)
  library(ComplexHeatmap)
  library(circlize)
  library(RColorBrewer)
  library(igraph)
  library(patchwork)
  library(cowplot)
  library(ggpubr)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Colors
```{r}
colors <- c(
  `Oligodendrocytes` = "royalblue3",
  `Microglia` = "orchid3",
  `OPCs` = "mediumpurple1",
  `Astrocytes` = "cornflowerblue",
  `Excitatory.Neurons` = "slateblue3",
  `Inhibitory.Neurons` = "darkslategray3",
  `Pericytes` = "deepskyblue3",
  `Endothelial cells` = "darkolivegreen4",
  `Meningeal cells` = "seagreen4",
  `Ependymal cells` = "lightcoral",
  `Fibroblasts` = "olivedrab3",
  `RGCs` = "violetred2"
)
```

# Load Data
* Filtered target gene DESeq2 output
* MultiNicheNet output
* AD risk gene list
* Network properties
* Unfiltered DESeq2 outputs for receivers
```{r}
# Target gene expression info from DESeq2
all_gex <- read_rds(here(
  "results",
  "final_outputs",
  "04_dea",
  "filtered_dea_targets.rds"
))

# MultiNicheNet object
output <- read_rds(here("data", "ccc", "multinichenet_output.rds"))

# AD gene list
ad_genes_mm <- read_rds(here("doc", "ad_genes_mm.rds"))

# Network properties
network_properties_6m <- read_rds(here(
  "results",
  "intermediate_outputs",
  "05_grn",
  "signaling_grn_properties_6m.rds"))

network_properties_12m <- read_rds(here(
  "results",
  "intermediate_outputs",
  "05_grn",
  "signaling_grn_properties_12m.rds"))

# Unfiltered DESeq2 outputs for receivers
excitatory_neurons_12m <- read_csv(here(
  "results",
  "intermediate_outputs",
  "03_dea",
  "excitatory_neurons_group_12mAD_vs_12mWT_all_genes.csv"
)) %>%
  mutate(cell_type = "Excitatory Neurons") %>%
  mutate(time_point = "12m")

inhibitory_neurons_12m <- read_csv(here(
  "results",
  "intermediate_outputs",
  "03_dea",
  "inhibitory_neurons_group_12mAD_vs_12mWT_all_genes.csv"
)) %>%
  mutate(cell_type = "Inhibitory Neurons") %>%
  mutate(time_point = "12m")
```

# Filter MultiNicheNet output
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# vector of groups of interest
group_oi <- c("X6mAD", "X6mWT", "X12mAD", "X12mWT")

# Filter output using above criteria and add time point column
output_filt <- filter_nichenet(output)
```

# Filter network properties
* We are interested in the signaling mediators
  * In the network, we deemed any outgoing edge from the receptor to another node constitutes a mediator
  * Therefore, we are grabbing the neighbors of our AD-associated interactions here
```{r}
# 6 month properties
properties_6m_filt <- get_mediators(network_properties_6m)

meds_in_6m <- properties_6m_filt$meds_in

meds_ex_6m <- properties_6m_filt$meds_ex

# 12 month properties
properties_12m_filt <- get_mediators(network_properties_12m)

meds_in_12m <- properties_12m_filt$meds_in

meds_ex_12m <- properties_12m_filt$meds_ex
```

# Filter target DEA info by AD-associated genes
```{r}
targets_ad <- all_gex %>%
  filter(gene %in% ad_genes_mm)

# filter by significance
sig_ad_targets <- targets_ad %>%
  filter(padj < 0.05)
```

# Prepare df for plotting panel A
```{r}
# Filter MultiNicheNet object by AD genes
ad_lrt <- output_filt %>%
  select(sender, receiver, ligand, receptor, target, group) %>%
  filter(target %in% sig_ad_targets$gene) %>%
  unique()

# Add asterisks to AD-associated genes for ligands and receptors
ad_lrt$ligand <- ifelse(ad_lrt$ligand %in% ad_genes_mm,
                        paste0(ad_lrt$ligand, "*"),
                        paste(ad_lrt$ligand))

ad_lrt$receptor <- ifelse(ad_lrt$receptor %in% ad_genes_mm,
                          paste0(ad_lrt$receptor, "*"),
                          paste(ad_lrt$receptor))

# Factorize columns of interest (for plotting purposes)
vec <- c("ligand", "receptor", "target")

ad_lrt[vec] <- lapply(ad_lrt[vec], function(x) factor(x, levels = unique(x)))

# Reorder columns of interest by the sender to aid with legibility
ad_lrt$ligand <- reorder(
  ad_lrt$ligand,
  desc(ad_lrt$sender)
)

ad_lrt$receptor <- reorder(
  ad_lrt$receptor,
  desc(ad_lrt$sender)
)
```

# Figure 5A
* Alluvial plot of LRs that have an AD-associated target gene
```{r}
# Filter for unique LRs only
lr_all <- ad_lrt %>%
  select(sender, ligand, receptor) %>%
  unique()

# Plot
alluvial_lr_all <- ggplot(lr_all,
  aes(axis1 = ligand, axis2 = receptor),
  label = stratum
) +
  scale_x_discrete(limits = c("Ligand", "Receptor")) +
  xlab("Type") +
  geom_alluvium() +
  stat_alluvium(lode.guidance = "frontback") +
  geom_stratum(aes(fill = sender)) +
  geom_label(
    stat = "stratum",
    fill = "white",
    aes(label = after_stat(stratum)),
    size = 3,
    fontface = "bold"
  ) +
  geom_flow(aes(fill = sender), stat = "alluvium", color = "black") +
  scale_fill_manual(values = colors) +
  scale_x_discrete(limits = c("Ligand", "Receptor")) +
  theme_void() +
  theme(
    axis.text.x = element_text(
      vjust = 2.5,
      face = "bold",
      size = 12,
      hjust = 0.5
    ),
    legend.position = c(0.9, 0.5),
    legend.margin = margin(0, 0, 0, 0.5, "cm"),
    legend.text = element_text(face = "bold", size = 10),
    legend.title = element_text(face = "bold", size = 12),
    plot.background = element_rect(fill = "white", size = 0),
    plot.margin = grid::unit(c(-0.4, 0.3, 0, -2.5), "cm"),
    axis.ticks = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    panel.spacing = element_blank(),
    axis.title.x = element_blank()
  ) +
  guides(fill = guide_legend(title = "Sender"))

alluvial_lr_all
```

# Save panel A
```{r}
ggsave(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "alluvial_lr_all.png"
  ),
  alluvial_lr_all,
  dpi = 300,
  width = 8,
  height = 8,
  bg = "#ffffff"
)
```

# Prepare inputs for panel B
```{r}
# Grab gene expression info for mediators by group from DESeq2 outputs
meds_inhib_12m <- inhibitory_neurons_12m %>%
  filter(gene %in% meds_in_12m)

meds_excit_12m <- excitatory_neurons_12m %>%
  filter(gene %in% meds_ex_12m)

# combine into 1 df
meds_all <- rbind(meds_inhib_12m, meds_excit_12m)
```

# Prepare mediator heatmap matrix
```{r}
# Pivot mediator gene expression df by group
dea_df <- meds_all %>%
  select(gene, log2FoldChange, cell_type, time_point) %>%
  pivot_wider(
    id_cols = gene,
    names_from = c(cell_type, time_point),
    values_from = log2FoldChange
  ) %>%
  column_to_rownames(var = "gene")

# Replace NAs with 0
dea_df[is.na(dea_df)] <- 0

# Create second df which includes the p-values
dea_df_pvalue <- meds_all %>%
  select(gene, padj, cell_type, time_point) %>%
  pivot_wider(
    id_cols = gene,
    names_from = c(cell_type, time_point),
    values_from = padj
  ) %>%
  column_to_rownames(var = "gene")

# Replace NAs with not significant (NS) and convert to a matrix
dea_df_pvalue[is.na(dea_df_pvalue)] <- "NS"
mat_dea_pvalue <- as.matrix(dea_df_pvalue)

# Prepare heatmap annotation
meta <- as.data.frame(colnames(dea_df))
rownames(meta) <- meta$`colnames(dea_df)`

meta$cell_type <- sub("_.*", "", colnames(dea_df))
meta$timepoint <- sub(".*_", "", colnames(dea_df))

meta <- meta %>%
  select(cell_type, timepoint)

# Set annotation colors
anno_cols <- list(
  "cell_type" = c(
    "Excitatory Neurons" = "slateblue3",
    "Inhibitory Neurons" = "darkslategray3"
  ),
  "timepoint" = c(
    "6m" = "#6A3D9A",
    "12m" = "#CAB2D6"
  )
)

# Generate annotation
anno <- HeatmapAnnotation(
  df = meta,
  show_annotation_name = FALSE,
  col = anno_cols,
  annotation_legend_param = list(
    cell_type = list(
      title = "Cell Type",
      direction = "horizontal",
      labels_gp = gpar(fontface = "bold", fontsize = 10),
      title_gp = gpar(fontface = "bold", fontsize = 12)
    ),
    timepoint = list(
      title = "Time Point",
      direction = "horizontal",
      labels_gp = gpar(fontface = "bold", fontsize = 10),
      title_gp = gpar(fontface = "bold", fontsize = 12)
    )
  )
)

# Convert df to matrix
mat_dea <- as.matrix(dea_df)
```

# Prepare AD target gene heatmap matrix
```{r}
# Remove the period in cell type names
sig_ad_targets$receiver <- gsub("\\.", " ", sig_ad_targets$receiver)

# Pivot target gene expression df by group
target_df <- sig_ad_targets %>%
  select(gene, log2FoldChange, receiver, timepoint) %>%
  pivot_wider(
    id_cols = gene,
    names_from = c(receiver, timepoint),
    values_from = log2FoldChange
  ) %>%
  column_to_rownames(var = "gene")

# Grab column order from mediator df
col_order <- colnames(dea_df)

# Reorder columns
target_df <- target_df[, col_order]

# Replace NAs with 0
target_df[is.na(target_df)] <- 0

# Convert df to matrix for plotting
mat_target <- as.matrix(target_df)

# Create second df with p-values
target_df_pvalue <- sig_ad_targets %>%
  select(gene, padj, receiver, timepoint) %>%
  pivot_wider(
    id_cols = gene,
    names_from = c(receiver, timepoint),
    values_from = padj
  ) %>%
  column_to_rownames(var = "gene")

# As before, re-order columns to match mediator df
target_df_pvalue <- target_df_pvalue[, col_order]

# Replace NAs
target_df_pvalue[is.na(target_df_pvalue)] <- "NS"

# Convert to matrix
mat_target_pvalue <- as.matrix(target_df_pvalue)
```

# Figure 5B
* Two heatmaps combined into 1
  * Top: AD-risk gene targets gex
  * Bottom: Signaling mediators gex
```{r}
# Set heatmap colors
cols <- colorRamp2(c(-1, -0.5, 0, 0.5, 1),
                   c("darkblue", "lightblue", "grey95", "tomato", "darkred"))

# AD-associated target gene heatmap
set.seed(42)
target_heatmap <- Heatmap(
  mat_target,
  col = cols,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if(mat_target_pvalue[i, j] < 0.05) {
      grid.text("*", x, y,
                gp = gpar(fontsize = 14,
                          fontface = "bold",
                          col = "white"))
    } else {
      grid.text("", x, y)
    }},
  show_column_names = FALSE,
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_title = "AD-associated Targets",
  row_title_gp = gpar(fontface = "bold", fontsize = 12),
  row_names_gp = gpar(fontface = "bold", fontsize = 10),
  heatmap_legend_param = list(
    direction = "horizontal",
    title = "AD-WT Log2FC",
    labels_gp = gpar(fontface = "bold", fontsize = 10),
    title_gp = gpar(fontface = "bold", fontsize = 12)
  )
)

# Mediator heatmap
set.seed(42)
mediator_heatmap <- Heatmap(
  mat_dea,
  col = cols,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if(mat_dea_pvalue[i, j] < 0.05) {
      grid.text("*", x, y,
                gp = gpar(fontsize = 14,
                          fontface = "bold",
                          col = "black"))
    } else {
      grid.text("", x, y)
    }
  },
  show_column_names = FALSE,
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_title = "Signaling Mediators of AD-associated LRTs",
  row_title_gp = gpar(fontface = "bold", fontsize = 12),
  row_names_gp = gpar(fontface = "bold", fontsize = 10),
  show_heatmap_legend = FALSE
)

# Combine the annotation with the heatmaps
heatmap_ls <- anno %v% target_heatmap %v% mediator_heatmap

# Plot combined heatmaps
draw(heatmap_ls)
```

# Save 5B
```{r}
# Grab combined heatmap with legends to save
heatmaps <- grid.grabExpr(
  draw(heatmap_ls,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legend = TRUE
  )
)

# save
ggsave(
  here(
    "results",
    "final_outputs",
    "03_ccc",
    "ad_target_mediators_heatmap.png"
  ),
  heatmaps,
  dpi = 300,
  width = 8,
  height = 10,
  bg = "#ffffff"
)
```

# Compile Figure 5
```{r}
fig5 <- ggarrange(
  alluvial_lr_all,
  heatmaps,
  align = "hv",
  ncol = 2,
  labels = c("A", "B")
)
```

# Save figure 5
```{r}
ggsave(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "figure5.png"
  ),
  fig5,
  dpi = 300,
  width = 13,
  height = 8,
  bg = "#ffffff"
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          here_1.0.1            ggpubr_0.6.0         
 [5] cowplot_1.1.1         patchwork_1.1.2       igraph_1.4.2          RColorBrewer_1.1-3   
 [9] circlize_0.4.15       ComplexHeatmap_2.14.0 ggalluvial_0.12.5     lubridate_1.9.2      
[13] forcats_1.0.0         stringr_1.5.0         dplyr_1.1.1           purrr_1.0.1          
[17] readr_2.1.4           tidyr_1.3.0           tibble_3.2.1          ggplot2_3.4.2        
[21] tidyverse_2.0.0      

loaded via a namespace (and not attached):
 [1] matrixStats_0.63.0  bit64_4.0.5         doParallel_1.0.17   rprojroot_2.0.3    
 [5] R.cache_0.16.0      tools_4.2.3         backports_1.4.1     utf8_1.2.3         
 [9] R6_2.5.1            BiocGenerics_0.44.0 lazyeval_0.2.2      colorspace_2.1-0   
[13] GetoptLong_1.0.5    withr_2.5.0         tidyselect_1.2.0    processx_3.8.1     
[17] bit_4.0.5           compiler_4.2.3      textshaping_0.3.6   cli_3.6.1          
[21] Cairo_1.6-0         xml2_1.3.3          desc_1.4.2          labeling_0.4.2     
[25] scales_1.2.1        callr_3.7.3         systemfonts_1.0.4   digest_0.6.31      
[29] rmarkdown_2.21      R.utils_2.12.2      pkgconfig_2.0.3     htmltools_0.5.5    
[33] fastmap_1.1.1       rlang_1.1.0         GlobalOptions_0.1.2 rstudioapi_0.14    
[37] farver_2.1.1        shape_1.4.6         generics_0.1.3      vroom_1.6.1        
[41] car_3.1-2           R.oo_1.25.0         magrittr_2.0.3      munsell_0.5.0      
[45] S4Vectors_0.36.2    fansi_1.0.4         abind_1.4-5         lifecycle_1.0.3    
[49] R.methodsS3_1.8.2   stringi_1.7.12      yaml_2.3.7          carData_3.0-5      
[53] parallel_4.2.3      crayon_1.5.2        hms_1.1.3           knitr_1.42         
[57] ps_1.7.5            pillar_1.9.0        rjson_0.2.21        ggsignif_0.6.4     
[61] codetools_0.2-19    stats4_4.2.3        glue_1.6.2          evaluate_0.20      
[65] rex_1.2.1           remotes_2.4.2       png_0.1-8           vctrs_0.6.2        
[69] tzdb_0.3.0          foreach_1.5.2       gtable_0.3.3        clue_0.3-64        
[73] xfun_0.38           broom_1.0.4         cyclocomp_1.1.0     rstatix_0.7.2      
[77] ragg_1.2.5          iterators_1.0.14    IRanges_2.32.0      cluster_2.1.4      
[81] timechange_0.2.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_5.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_5.Rmd"
))
```
