---
title: "figure_6"
author: "Tabea M. Soelter"
date: "2024-02-19"
output: html_document
---
**Plotting Manuscript Figure 6 and Supplementary Figure 6**

__Goal__: This is a script for generating figure 6 of my manuscript. We are also saving supp figure 6.

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*AD Gene List*
* Name: ad_genes_mm.rds
* Location: doc/
*12m Network properties*
* Name: signaling_grn_properties_12m.rds
* Location: results/intermediate_outputs/05_grn/
*AD enriched gene targeting*
* Name: filtered_targeting_ad.rds
* Location: results/intermediate_outputs/07_gene_targeting/
*TF activity*
* Names: 
  1. excitatory_tf_activity.rds
  2. inhibitory_tf_activity.rds
* Location: results/final_outputs/05_bio_activity/

__Analysis Plan__:
* Load necessary packages
* Load data
* Compile 12m mediators
* Plot individual panels:
  * A: Heatmap of differential gene targeting of signaling mediators
        * Only include mediators who are in top or bottom quartile
        * They are the most up- or down-regulated
  * B: Heatmap of TF activity of mediators that are TFs
* Compile figure 6
* Save figure 6
* Plot figure S6:
  * Dotplot of global diff targeting to give context to mediator targeting
* Save figure S6

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# set seed
set.seed(42)

# load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggalluvial)
  library(ComplexHeatmap)
  library(circlize)
  library(RColorBrewer)
  library(igraph)
  library(patchwork)
  library(cowplot)
  library(ggpubr)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
# AD gene list
ad_genes <- read_rds(here("doc", "ad_genes_mm.rds"))

# Load signaling network properties
network_properties_12m <- read_rds(here(
  "results",
  "intermediate_outputs",
  "05_grn",
  "signaling_grn_properties_12m.rds"
))

# AD enriched gene targeting
filtered_ad <- readRDS(here(
  "results",
  "intermediate_outputs",
  "07_gene_targeting",
  "filtered_targeting_ad.rds"
))

# TF activity
tf_acts_ex <- readRDS(here(
  "results",
  "final_outputs",
  "05_bio_activity",
  "excitatory_tf_activity.rds"
)) %>%
  mutate(cell_type = "Excitatory Neurons")

tf_acts_in <- readRDS(here(
  "results",
  "final_outputs",
  "05_bio_activity",
  "inhibitory_tf_activity.rds"
)) %>%
  mutate(cell_type = "Inhibitory Neurons")
```

## Figure 6A
# Create vector of 12m mediators
```{r}
properties_12m_filt <- get_mediators(network_properties_12m)

meds_all <- c(properties_12m_filt$meds_in, properties_12m_filt$meds_ex)
```

# Filter AD enriched scores by mediators
```{r}
mediator_df <- filtered_ad %>%
  filter(gene %in% meds_all)
```

# Plot Figure 6A
```{r}
# prepare matrix for plotting
meds_df <- pivot_wider(mediator_df,
  names_from = gene,
  values_from = targeting_score
) %>%
  column_to_rownames("cell_type")

# Replace NAs with 0 (= not enriched genes, but needed for plotting)
meds_df[is.na(meds_df)] <- 0

# Convert to matrix
meds_mat <- as.matrix(meds_df)

# prepare top annotation
cell_type <- c("Excitatory Neurons", "Inhibitory Neurons")
meta <- cell_type %>% as.data.frame()
rownames(meta) <- c("excitatory_neurons", "inhibitory_neurons")
colnames(meta) <- "cell_type"

# Set annotation colors
anno_cols <- list(
  "cell_type" = c(
    "Excitatory Neurons" = "slateblue3",
    "Inhibitory Neurons" = "darkslategray3"
  )
)

# Set heatmap colors and breaks
cols <- colorRamp2(c(-250, 0, 250), rev(brewer.pal(n = 3, name = "RdBu")))

# Make heatmap annotation
anno <- HeatmapAnnotation(
  df = meta,
  show_annotation_name = FALSE,
  col = anno_cols,
  annotation_legend_param = list(
    cell_type = list(
      title = "Receiver",
      direction = "horizontal",
      labels_gp = gpar(fontface = "bold")
    )
  )
)

# Plot heatmap
gene_targeting_heatmap <- Heatmap(t(meds_mat),
  col = cols,
  top_annotation = anno,
  show_column_names = FALSE,
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_title = "AD-enriched Signaling Mediators",
  row_title_gp = gpar(
    fontface = "bold",
    fontsize = 12
  ),
  row_names_gp = gpar(fontface = "bold"),
  heatmap_legend_param = list(
    direction = "vertical",
    title = "Gene Targeting Score",
    labels_gp = gpar(fontface = "bold"),
    labels = c("-400", "-200", "NE", "200", "400")
  )
)

draw(gene_targeting_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legend = TRUE
)
```

# Save panel 6A
```{r}
# grab final figure panel
panel_a <- grid.grabExpr(
  draw(
    gene_targeting_heatmap,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legend = TRUE
  )
)

# save
ggsave(
  here(
    "results",
    "final_outputs",
    "06_gene_targeting",
    "ad_enriched_mediator_gene_targeting.png"
  ),
  panel_a,
  dpi = 300,
  width = 7,
  height = 5,
  bg = "#ffffff"
)
```

## Figure 6B
# Combine TF activity outputs
```{r}
tf_acts_all <- rbind(tf_acts_ex, tf_acts_in)
```

# Filter TF activity by mediators
```{r}
tf_acts_all_filt <- tf_acts_all %>%
  filter(source %in% meds_all) %>%
  filter(condition == "12m")

dim(tf_acts_all_filt) # 16 6
```

# Plot Figure 6B
```{r}
# prepare matrix for plotting
mat <- tf_acts_all_filt %>%
  select(source, score, cell_type) %>%
  pivot_wider(
    names_from = "source",
    values_from = "score"
  ) %>%
  column_to_rownames(var = "cell_type") %>%
  as.matrix()

# prepare top annotation
cell_type <- c("Excitatory Neurons", "Inhibitory Neurons")
meta <- cell_type %>% as.data.frame()
rownames(meta) <- c("Excitatory Neurons", "Inhibitory Neurons")
colnames(meta) <- "cell_type"

# Set heatmap colors and breaks
cols <- colorRamp2(c(-2, 0, 2), brewer.pal(n = 3, name = "PRGn"))

# Plot tf activity heatmap
tf_activity_heatmap <- Heatmap(t(mat),
  col = cols,
  top_annotation = anno,
  show_column_names = FALSE,
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_title = "Transcription Factors",
  row_title_gp = gpar(
    fontface = "bold",
    fontsize = 12
  ),
  row_names_gp = gpar(fontface = "bold"),
  heatmap_legend_param = list(
    direction = "vertical",
    title = "TF Activity",
    labels_gp = gpar(fontface = "bold")
  )
)

draw(tf_activity_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legend = TRUE
)
```

# Save panel 6B
```{r}
# grab figure panel
panel_b <- grid.grabExpr(
  draw(
    tf_activity_heatmap,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legend = TRUE
  )
)

# save
ggsave(
  here(
    "results",
    "final_outputs",
    "06_gene_targeting",
    "tf_activity_heatmap_sig_meds.png"
  ),
  panel_b,
  dpi = 300,
  width = 7,
  height = 5,
  bg = "#ffffff"
)
```

# Compile Figure 6
```{r}
fig6 <- plot_grid(panel_a,
  panel_b,
  labels = c("A", "B")
)

fig6
```

# Save figure 6
```{r}
ggsave(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "figure6.png"
  ),
  fig6,
  dpi = 300,
  width = 12,
  height = 5,
  bg = "#ffffff"
)
```

# Figure S6
* Global gene targeting in receivers
  * Will show how our mediators stack up against
```{r}
filtered_ad_wide <- pivot_wider(filtered_ad,
  values_from = "targeting_score",
  names_from = "cell_type"
)

# replace all negatives and NA with 0 for positive scores
pos <- filtered_ad_wide %>%
  mutate_at(vars(-gene), ~ replace(., is.na(.) | . < 0, 0))
pos_non_zero_counts <- sapply(pos[-1], function(x) sum(x != 0))
pos_non_zero_counts

# sum diff targeting by cell type
pos <- colSums(pos[, -1])

## replace all positives and NA with 0 for negative scores
neg <- filtered_ad_wide %>%
  mutate_at(vars(-gene), ~ replace(., is.na(.) | . > 0, 0))
neg_non_zero_counts <- sapply(neg[-1], function(x) sum(x != 0))
neg_non_zero_counts

## sum diff targeting by cell type
neg <- colSums(neg[, -1])

# Combine positive and negative
data <- as.data.frame(cbind(pos, neg))
rownames(data) <- c("Excitatory Neurons", "Inhibitory Neurons")
data$cell_type <- rownames(data)
data$pos_sum <- pos_non_zero_counts
data$neg_sum <- neg_non_zero_counts

data <- data[order(data$pos, decreasing = TRUE), ]
```
excitatory_neurons inhibitory_neurons 
              6317               4020 
excitatory_neurons inhibitory_neurons 
              3559               5856 

# Plot Figure S6
```{r}
# formatting data for plotting
data_plot <- pivot_longer(data[, c(1:3)], cols = c(1, 2), values_to = "sum")
data_label <- pivot_longer(data[, c(4, 5)], cols = c(1, 2), values_to = "sum")
data_plot$num <- data_label$sum

# normalize data by number of genes with enriched targeting
data_plot$norm <- data_plot$sum / data_plot$num

# set plot order
data_plot <- arrange(data_plot, desc(norm))
data_plot$order <- seq(1:4)

# plot
dotplot_global_targeting <- ggplot(
  data_plot,
  aes(
    x = norm,
    y = reorder(cell_type, norm)
  )
) +
  geom_point(shape = 21, aes(fill = name, color = name, size = num)) +
  theme_bw() +
  scale_color_manual(
    values = c(
      "black",
      "black"
    ),
    guide = FALSE
  ) +
  scale_fill_manual(
    values = c(
      "blue",
      "tomato"
    ),
    labels = c(
      "Decreased in 3xTg-AD",
      "Increased in 3xTg-AD"
    )
  ) +
  labs(fill = "Differential Gene Targeting") +
  theme(legend.position = "right") +
  ylab("Cell Type") +
  xlab("Total differential targeting score") +
  theme(axis.text.x = element_text(
    face = "bold",
    color = "black",
    size = 10
  )) +
  theme(axis.title.x = element_text(
    face = "bold",
    color = "black",
    size = 12
  )) +
  theme(axis.title.y = element_text(
    face = "bold",
    color = "black",
    size = 12
  )) +
  geom_text(aes(label = num), vjust = -1, fontface = "bold") +
  theme(axis.text.y = element_text(
    face = "bold",
    color = "black",
    size = 10
  )) +
  guides(size = guide_legend(title = "Number of Genes")) +
  theme(text = element_text(face = "bold")) +
  xlim(c(-300, 300))

dotplot_global_targeting
```

# Save Figure S5
```{r}
ggsave(
  here(
    "results",
    "final_outputs",
    "01_figures",
    "supp_figure6.png"
  ),
  dotplot_global_targeting,
  dpi = 300,
  width = 8,
  height = 5,
  bg = "#ffffff"
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.4 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          here_1.0.1           
 [4] ggpubr_0.6.0          cowplot_1.1.1         patchwork_1.1.2      
 [7] igraph_1.4.2          RColorBrewer_1.1-3    circlize_0.4.15      
[10] ComplexHeatmap_2.14.0 ggalluvial_0.12.5     lubridate_1.9.2      
[13] forcats_1.0.0         stringr_1.5.0         dplyr_1.1.1          
[16] purrr_1.0.1           readr_2.1.4           tidyr_1.3.0          
[19] tibble_3.2.1          ggplot2_3.4.2         tidyverse_2.0.0      

loaded via a namespace (and not attached):
 [1] matrixStats_0.63.0  doParallel_1.0.17   rprojroot_2.0.3     R.cache_0.16.0     
 [5] tools_4.2.3         backports_1.4.1     utf8_1.2.3          R6_2.5.1           
 [9] BiocGenerics_0.44.0 lazyeval_0.2.2      colorspace_2.1-0    GetoptLong_1.0.5   
[13] withr_2.5.0         tidyselect_1.2.0    processx_3.8.1      compiler_4.2.3     
[17] textshaping_0.3.6   cli_3.6.1           Cairo_1.6-0         xml2_1.3.3         
[21] desc_1.4.2          labeling_0.4.2      scales_1.2.1        callr_3.7.3        
[25] systemfonts_1.0.4   digest_0.6.31       rmarkdown_2.21      R.utils_2.12.2     
[29] pkgconfig_2.0.3     htmltools_0.5.5     fastmap_1.1.1       rlang_1.1.0        
[33] GlobalOptions_0.1.2 rstudioapi_0.14     farver_2.1.1        shape_1.4.6        
[37] generics_0.1.3      car_3.1-2           R.oo_1.25.0         magrittr_2.0.3     
[41] munsell_0.5.0       S4Vectors_0.36.2    fansi_1.0.4         abind_1.4-5        
[45] lifecycle_1.0.3     R.methodsS3_1.8.2   stringi_1.7.12      yaml_2.3.7         
[49] carData_3.0-5       parallel_4.2.3      crayon_1.5.2        hms_1.1.3          
[53] knitr_1.42          ps_1.7.5            pillar_1.9.0        rjson_0.2.21       
[57] ggsignif_0.6.4      codetools_0.2-19    stats4_4.2.3        glue_1.6.2         
[61] evaluate_0.20       rex_1.2.1           remotes_2.4.2       png_0.1-8          
[65] vctrs_0.6.2         tzdb_0.3.0          foreach_1.5.2       gtable_0.3.3       
[69] clue_0.3-64         xfun_0.38           broom_1.0.4         cyclocomp_1.1.0    
[73] rstatix_0.7.2       ragg_1.2.5          iterators_1.0.14    IRanges_2.32.0     
[77] cluster_2.1.4       timechange_0.2.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "figures",
  "figure_6.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "figures",
  "figure_6.Rmd"
))
```
