---
title: "01_3xtgad_seurat_preprocessing"
author: "Tabea M. Soelter"
date: "2023-05-08"
output: html_document
---
**Pre-processing of 3xTgAD mouse data**

__Goal__: Process 3xTgAD (snRNA-seq) data for downstream analyses. The data consists of 2 time points (6 and 12 months), 2 conditions (AD, CTRL), and 1 sex (female).
  
__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.0 
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 65GB per CPU

__Data__:
* Name: N/A
* Location: /data/user/tsoelter/projects/230418_TS_AgingCCC/data/CellRangerCounts/
 
__Analysis Plan__:
* Load necessary packages 
* Load data 
* Create seurat object
* Perform quality control and filtering
* Integration
* Save integrated seurat object

__Analysis__:
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(Seurat)
  library(tidyverse)
  library(harmony)
  library(stringr)
  library(scCustomize)
  library(biomaRt)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Create seurat data directory
```{r}
if (!dir.exists(here("data", "seurat"))) dir.create(here("data", "seurat"))
```

# Create seurat object
```{r}
merged_seurat <- make_seurat_object(here(
  "data",
  "CellRangerCounts",
  "post_soupX"
))
```

# Calculate QC metrics
```{r}
merged_seurat <- calculate_qc(merged_seurat)
```

# Format metadata
```{r}
metadata <- format_metadata(merged_seurat)
dim(metadata) # 185643 9

# Add metadata to merged_seurat
merged_seurat@meta.data <- metadata
```

# Plot QC metrics
```{r warning=FALSE}
pdf(file = here("results", "intermediate_outputs", "02_seurat", "qc_plots.pdf"))
plot_qc(metadata)
dev.off()
```

# Filtering 
* Performing cell-level and gene-level filtering on the data. 
* Cell-level filtering is performed based on the plotted QC metrics (above).
* Gene-level filtering aims at removing count values = 0, since they can skew average gene expression
measurements.
  * Discussed in the following review: https://genomebiology.biomedcentral.com/articles/10.1186/s13059-022-02601-5)
```{r}
# Cell-level filtering
sub_seurat <- subset(merged_seurat,
  subset =
    nGene > 500 &
      nGene < 10000 &
      mitoRatio < 0.05 &
      log10GenesPerUMI > 0.80
)

# Gene-level filtering
counts <- GetAssayData(sub_seurat, slot = "counts")

nonzero <- counts > 0

keep_genes <- rowSums(nonzero) >= 10

filtered_counts <- counts[keep_genes, ]

filtered_seurat <- CreateSeuratObject(filtered_counts,
  meta.data = sub_seurat@meta.data
)
```

# Save filtered seurat object
```{r}
# create directory if needed
filepath <- here("data", "seurat")
if (!dir.exists(filepath)) dir.create(filepath)
saveRDS(filtered_seurat, file = paste0(filepath, "filtered_seurat.rds"))
```

# Plot QC metrics after filtering
```{r warning=FALSE}
metadata <- filtered_seurat@meta.data
dim(metadata) # 184858 12

# plot and save post-filter QC outputs
pdf(file = here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "qc_plots_filtered.pdf"
))
plot_qc(metadata)
dev.off()
```

# Determine cell cycle effects
* Since the cell cycle genes included in Seurat only have human IDs, I have to convert them to mouse genes for this data.
* To do so, I am using a recommendation described here:
  * https://github.com/satijalab/seurat/issues/2493
  * https://www.r-bloggers.com/2016/10/converting-mouse-to-human-gene-names-with-biomart-package/
```{r warning=FALSE}
pdf(file = here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "cellcycle_pca.pdf"
))

filtered_seurat <- cell_cycle_effects(filtered_seurat)

dev.off()
```

# Normalization and integration
* We are using harmony as it has been shown to remove technical variation while preserving biological variation.
* We will have to provide the PCs from the elbow plot above (1:20)
```{r warning=FALSE}
filtered_seurat <- harmony_integration(filtered_seurat, dims = 1:20)
# harmony converged after 10 iterations
```

# Save integrated seurat object
```{r}
saveRDS(filtered_seurat, file = here("data", "seurat", "integrated_seurat.rds"))
```

# Clustering
* Since I am using harmony for integration, I will have to ensure to use the RNA assay and the harmony reduction for everything downstream.
* I am using Leiden for clustering in my find_clusters function. 
```{r warning=FALSE}
# check active assay (needs to be RNA)
filtered_seurat@active.assay

# vector of desired resolutions
# resolutions <- c(
#   0.5, 0.6, 0.7, 0.8, 0.9,
#   1.0, 1.1, 1.2, 1.3, 1.4, 1.5
# )
# selected resolution
resolutions <- 0.7

# clustering of resolutions (answer no to miniconda installation)
filtered_seurat <- find_clusters(filtered_seurat,
  dims = 1:20,
  reduction = "harmony",
  resolutions = resolutions
)
```

# Save clustered seurat object
```{r}
saveRDS(filtered_seurat, file = here("data", "seurat", "clustered_seurat.rds"))
```

# Determine resolution
```{r}
# Plot resolutions between 0.1 and 0.9
tree <- clustree(filtered_seurat@meta.data, prefix = "RNA_snn_res.0.")

# Resolutions between 1 and 1.9
tree2 <- clustree(filtered_seurat@meta.data, prefix = "RNA_snn_res.1.")

pdf(file = here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "clustree_umap.pdf"
))

tree
tree2

# plotting UMAP for most stable resolution (0.7)
DimPlot(filtered_seurat,
  reduction = "umap_harmony",
  group.by = "RNA_snn_res.0.7"
)

dev.off()
```

# Find marker genes
```{r}
# Set the identity as Leiden with resolution 0.7
filtered_seurat <- SetIdent(filtered_seurat, value = "RNA_snn_res.0.7")

# Find marker genes
all_marker_genes <- FindAllMarkers(filtered_seurat,
  log2FC.threshold = 0.2,
  test.use = "wilcox",
  min.pct = 0.1,
  only.pos = TRUE,
  max.cells.per.ident = 50,
  assay = "RNA"
)

# Save marker genes
saveRDS(all_marker_genes, here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "all_marker_genes.rds"
))

# Look at the top 25 up-regulated genes
top25 <- all_marker_genes %>%
  group_by(cluster) %>%
  top_n(-25, p_val_adj)

dim(top25) # 1074 7
```

# Neuronal marker FeaturePlots
* As no markers from PanglaoDB indicated whether Neurons are excitatory or inhibitory, I am using feature plots and known canonical markers to identify neuronal clusters as ex or in.
* I was able to assign all neuron clusters as excitatory or inhibitory to most neuron clusters (cluster 4 excluded).
```{r}
pdf(file = here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "neurons_feature_plots.pdf"
))

# Slc17a7 (Excitatory Neurons) vs Gad2 (Inhibitory Neurons)
FeaturePlot(filtered_seurat, features = c("Slc17a7", "Gad2"), label = TRUE)
FeaturePlot(filtered_seurat, features = c("Slc17a7", "Gad2"), label = FALSE)

# Slc17a6 (Excitatory Neurons) vs Gad1 (Inhibitory Neurons)
FeaturePlot(filtered_seurat, features = c("Slc17a6", "Gad1"), label = TRUE)
FeaturePlot(filtered_seurat, features = c("Slc17a6", "Gad1"), label = FALSE)

dev.off()
```

# Finding markers of unresolved clusters
* Clusters unidentified are: 15, 22, 35, 38
```{r}
# vector of clusters of interest
identities <- c("15", "22", "35", "38")

# find marker genes across clusters of interest
top_markers <- find_markers(filtered_seurat,
  resolution = "RNA_snn_res.0.7",
  identities = identities,
  value = 15
)

dim(top_markers) # 60 6
```

# Assign cell types
```{r}
filtered_seurat <- RenameIdents(filtered_seurat,
  `1`  = "Astrocytes",
  `2`  = "Excitatory Neurons",
  `3`  = "Oligodendrocytes",
  `4`  = "Inhibitory Neurons",
  `5`  = "Inhibitory Neurons",
  `6`  = "Inhibitory Neurons",
  `7`  = "Excitatory Neurons",
  `8`  = "Oligodendrocytes",
  `9`  = "Excitatory Neurons",
  `10` = "Excitatory Neurons",
  `11` = "Excitatory Neurons",
  `12` = "OPCs",
  `13` = "Inhibitory Neurons",
  `14` = "Vasculature cells",
  `15` = "Ependymal cells",
  `16` = "Oligodendrocytes",
  `17` = "Excitatory Neurons",
  `18` = "Microglia",
  `19` = "Astrocytes",
  `20` = "Inhibitory Neurons",
  `21` = "Excitatory Neurons",
  `22` = "Unknown",
  `23` = "Inhibitory Neurons",
  `24` = "Inhibitory Neurons",
  `25` = "Inhibitory Neurons",
  `26` = "Excitatory Neurons",
  `27` = "Vasculature cells",
  `28` = "Inhibitory Neurons",
  `29` = "Astrocytes",
  `30` = "Excitatory Neurons",
  `31` = "Endothelial cells",
  `32` = "OPCs",
  `33` = "Inhibitory Neurons",
  `34` = "Excitatory Neurons",
  `35` = "Vasculature cells",
  `36` = "Excitatory Neurons",
  `37` = "Microglia",
  `38` = "Retinal ganglion cells",
  `39` = "OPCs",
  `40` = "Inhibitory Neurons",
  `41` = "Vasculature cells",
  `42` = "Excitatory Neurons"
)
```

# Plot final UMAP
```{r}
png(file = here("results", "final_outputs", "02_seurat", "UMAP_final.png"))
DimPlot_scCustom(filtered_seurat,
  figure_plot = TRUE,
  label = FALSE,
  ggplot_default_colors = TRUE
)
dev.off()
```

# Save processed seurat object
```{r}
saveRDS(filtered_seurat, here("data", "seurat", "processed_seurat.rds"))
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2        styler_1.9.1       here_1.0.1         biomaRt_2.54.1    
 [5] scCustomize_1.1.1  harmony_0.1.0      Rcpp_1.0.10        lubridate_1.9.2   
 [9] forcats_1.0.0      stringr_1.5.0      dplyr_1.1.1        purrr_1.0.1       
[13] readr_2.1.4        tidyr_1.3.0        tibble_3.2.1       ggplot2_3.4.2     
[17] tidyverse_2.0.0    SeuratObject_4.1.3 Seurat_4.3.0      

loaded via a namespace (and not attached):
  [1] utf8_1.2.3             R.utils_2.12.2         spatstat.explore_3.1-0
  [4] reticulate_1.28        tidyselect_1.2.0       RSQLite_2.3.1         
  [7] AnnotationDbi_1.60.2   htmlwidgets_1.6.2      grid_4.2.3            
 [10] Rtsne_0.16             munsell_0.5.0          codetools_0.2-19      
 [13] ica_1.0-3              future_1.32.0          miniUI_0.1.1.1        
 [16] withr_2.5.0            spatstat.random_3.1-4  colorspace_2.1-0      
 [19] progressr_0.13.0       Biobase_2.58.0         filelock_1.0.2        
 [22] knitr_1.42             rstudioapi_0.14        stats4_4.2.3          
 [25] ROCR_1.0-11            tensor_1.5             listenv_0.9.0         
 [28] labeling_0.4.2         GenomeInfoDbData_1.2.9 polyclip_1.10-4       
 [31] farver_2.1.1           bit64_4.0.5            rprojroot_2.0.3       
 [34] parallelly_1.35.0      vctrs_0.6.2            generics_0.1.3        
 [37] xfun_0.38              timechange_0.2.0       BiocFileCache_2.6.1   
 [40] R6_2.5.1               GenomeInfoDb_1.34.9    ggbeeswarm_0.7.1      
 [43] rex_1.2.1              bitops_1.0-7           spatstat.utils_3.0-2  
 [46] cachem_1.0.7           promises_1.2.0.1       scales_1.2.1          
 [49] beeswarm_0.4.0         gtable_0.3.3           globals_0.16.2        
 [52] processx_3.8.1         goftest_1.2-3          rlang_1.1.0           
 [55] cyclocomp_1.1.0        GlobalOptions_0.1.2    splines_4.2.3         
 [58] lazyeval_0.2.2         spatstat.geom_3.1-0    yaml_2.3.7            
 [61] reshape2_1.4.4         abind_1.4-5            httpuv_1.6.9          
 [64] tools_4.2.3            ellipsis_0.3.2         RColorBrewer_1.1-3    
 [67] BiocGenerics_0.44.0    ggridges_0.5.4         plyr_1.8.8            
 [70] progress_1.2.2         zlibbioc_1.44.0        RCurl_1.98-1.12       
 [73] ps_1.7.5               prettyunits_1.1.1      deldir_1.0-6          
 [76] pbapply_1.7-0          cowplot_1.1.1          S4Vectors_0.36.2      
 [79] zoo_1.8-12             ggrepel_0.9.3          cluster_2.1.4         
 [82] magrittr_2.0.3         data.table_1.14.8      scattermore_0.8       
 [85] circlize_0.4.15        lmtest_0.9-40          RANN_2.6.1            
 [88] fitdistrplus_1.1-8     R.cache_0.16.0         matrixStats_0.63.0    
 [91] hms_1.1.3              patchwork_1.1.2        mime_0.12             
 [94] evaluate_0.20          xtable_1.8-4           XML_3.99-0.14         
 [97] IRanges_2.32.0         gridExtra_2.3          shape_1.4.6           
[100] compiler_4.2.3         KernSmooth_2.23-20     crayon_1.5.2          
[103] R.oo_1.25.0            htmltools_0.5.5        mgcv_1.8-42           
[106] later_1.3.0            tzdb_0.3.0             ggprism_1.0.4         
[109] DBI_1.1.3              dbplyr_2.3.2           MASS_7.3-58.3         
[112] rappdirs_0.3.3         Matrix_1.5-4           cli_3.6.1             
[115] R.methodsS3_1.8.2      parallel_4.2.3         igraph_1.4.2          
[118] pkgconfig_2.0.3        sp_1.6-0               plotly_4.10.1         
[121] spatstat.sparse_3.0-1  xml2_1.3.3             paletteer_1.5.0       
[124] vipor_0.4.5            XVector_0.38.0         snakecase_0.11.0      
[127] callr_3.7.3            digest_0.6.31          sctransform_0.4.1     
[130] RcppAnnoy_0.0.20       janitor_2.2.0          spatstat.data_3.0-1   
[133] Biostrings_2.66.0      rmarkdown_2.21         leiden_0.4.3          
[136] uwot_0.1.14            curl_5.0.0             shiny_1.7.4           
[139] lifecycle_1.0.3        nlme_3.1-162           jsonlite_1.8.4        
[142] desc_1.4.2             viridisLite_0.4.1      fansi_1.0.4           
[145] pillar_1.9.0           lattice_0.21-8         ggrastr_1.0.1         
[148] KEGGREST_1.38.0        fastmap_1.1.1          httr_1.4.5            
[151] survival_3.5-5         remotes_2.4.2          glue_1.6.2            
[154] png_0.1-8              bit_4.0.5              stringi_1.7.12        
[157] rematch2_2.1.2         blob_1.2.4             memoise_2.0.1         
[160] irlba_2.3.5.1          future.apply_1.10.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60) 

# Time elapsed: 45 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "seurat_preprocessing",
  "01_3xtgad_seurat_preprocessing.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "seurat_preprocessing",
  "01_3xtgad_seurat_preprocessing.Rmd"
))
```
