---
title: "03_3xtgad_celltype_assignment"
author: "Tabea M. Soelter"
date: "2023-11-21"
output: html_document
---
**Cell-type assignment for 3xTgAD mouse data**

__Goal__: Assign cell types for the 3xTgAD (snRNA-seq) data for downstream analyses.
  
__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.0 
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 65GB per CPU

__Data__:
* Name: clustered_seurat.rds
* Location: /data/user/tsoelter/projects/230418_TS_AgingCCC/data/seurat/
 
__Analysis Plan__:
* Load necessary packages 
* Load data 
* Cluster at most stable resolution
* Find marker genes
* Assign cell types using marker genes and feature plots
* Save processed seurat object

__Analysis__:
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(Seurat)
  library(tidyverse)
  library(stringr)
  library(scCustomize)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Clustering
__NOTE: ONLY RUN THIS CHUNK IF YOU DID NOT RUN OPTIONAL 02_3xtgad_clustering.R/sh SCRIPTS__
__* Skip to line 75 if you did run the above scripts__
  * 02 scripts were needed to determine most stable resolution (0.7)
```{r eval=FALSE, warning=FALSE, include=FALSE}
# load integrated seurat object
filtered_seurat <- readRDS(here("data", "seurat", "integrated_seurat.rds"))

# check active assay (needs to be RNA)
filtered_seurat@active.assay

# clustering of most stable resolution (determined using clustree - 02 scripts)
filtered_seurat <- find_clusters(filtered_seurat,
  dims = 1:20,
  reduction = "harmony",
  resolutions = 0.7
)

# save clustered object
saveRDS(filtered_seurat, file = here("data", "seurat", "clustered_seurat.rds"))
```

# Load clustered seurat object
```{r}
filtered_seurat <- readRDS(here("data", "seurat", "clustered_seurat.rds"))
```

# Find marker genes
```{r}
# Set the identity as Leiden with resolution 0.7
filtered_seurat <- SetIdent(filtered_seurat, value = "RNA_snn_res.0.7")

# Find marker genes
all_marker_genes <- FindAllMarkers(filtered_seurat,
  log2FC.threshold = 0.2,
  test.use = "wilcox",
  min.pct = 0.1,
  only.pos = TRUE,
  max.cells.per.ident = 50,
  assay = "RNA"
)

# Save marker genes
saveRDS(all_marker_genes, here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "all_marker_genes.rds"
))

# Look at the top 25 up-regulated genes
top25 <- all_marker_genes %>%
  group_by(cluster) %>%
  top_n(-25, p_val_adj)

dim(top25) # 1000 7
```

# Neuronal marker FeaturePlots
* As no markers from PanglaoDB indicated whether Neurons are excitatory or inhibitory, I am using feature plots and known canonical markers to identify neuronal clusters as ex or in.
* I was able to assign all neuron clusters as excitatory or inhibitory to most neuron clusters (cluster 4 excluded).
```{r}
pdf(file = here(
  "results",
  "intermediate_outputs",
  "02_seurat",
  "neurons_feature_plots.pdf"
))

# Slc17a7 (Excitatory Neurons) vs Gad2 (Inhibitory Neurons)
FeaturePlot(filtered_seurat, features = c("Slc17a7", "Gad2"), label = TRUE)
FeaturePlot(filtered_seurat, features = c("Slc17a7", "Gad2"), label = FALSE)

# Slc17a6 (Excitatory Neurons) vs Gad1 (Inhibitory Neurons)
FeaturePlot(filtered_seurat, features = c("Slc17a6", "Gad1"), label = TRUE)
FeaturePlot(filtered_seurat, features = c("Slc17a6", "Gad1"), label = FALSE)

dev.off()
```

# Finding markers of unresolved clusters
* Clusters unidentified are: 15, 22, 35, 38
```{r}
# vector of clusters of interest
identities <- c("15", "22", "35", "38")

# find marker genes across clusters of interest
top_markers <- find_markers(filtered_seurat,
  resolution = "RNA_snn_res.0.7",
  identities = identities,
  value = 15
)

dim(top_markers) # 60 6
```

# Assign cell types
```{r}
filtered_seurat <- RenameIdents(filtered_seurat,
  `1`  = "Astrocytes",
  `2`  = "Excitatory Neurons",
  `3`  = "Oligodendrocytes",
  `4`  = "Inhibitory Neurons",
  `5`  = "Inhibitory Neurons",
  `6`  = "Inhibitory Neurons",
  `7`  = "Excitatory Neurons",
  `8`  = "Oligodendrocytes",
  `9`  = "Excitatory Neurons",
  `10` = "Excitatory Neurons",
  `11` = "Excitatory Neurons",
  `12` = "OPCs",
  `13` = "Inhibitory Neurons",
  `14` = "Vasculature cells",
  `15` = "Ependymal cells",
  `16` = "Oligodendrocytes",
  `17` = "Excitatory Neurons",
  `18` = "Microglia",
  `19` = "Astrocytes",
  `20` = "Inhibitory Neurons",
  `21` = "Excitatory Neurons",
  `22` = "Unknown",
  `23` = "Inhibitory Neurons",
  `24` = "Inhibitory Neurons",
  `25` = "Inhibitory Neurons",
  `26` = "Excitatory Neurons",
  `27` = "Vasculature cells",
  `28` = "Inhibitory Neurons",
  `29` = "Astrocytes",
  `30` = "Excitatory Neurons",
  `31` = "Endothelial cells",
  `32` = "OPCs",
  `33` = "Inhibitory Neurons",
  `34` = "Excitatory Neurons",
  `35` = "Vasculature cells",
  `36` = "Excitatory Neurons",
  `37` = "Microglia",
  `38` = "Retinal ganglion cells",
  `39` = "OPCs",
  `40` = "Inhibitory Neurons",
  `41` = "Vasculature cells",
  `42` = "Excitatory Neurons"
)
```

# Plot final UMAP
```{r}
png(file = here("results", "final_outputs", "02_seurat", "UMAP_final.png"))
DimPlot(filtered_seurat)
dev.off()
```

# Save processed seurat object
```{r}
saveRDS(filtered_seurat, here("data", "seurat", "processed_seurat.rds"))
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2        styler_1.9.1       here_1.0.1         scCustomize_1.1.1 
 [5] lubridate_1.9.2    forcats_1.0.0      stringr_1.5.0      dplyr_1.1.1       
 [9] purrr_1.0.1        readr_2.1.4        tidyr_1.3.0        tibble_3.2.1      
[13] ggplot2_3.4.2      tidyverse_2.0.0    SeuratObject_4.1.3 Seurat_4.3.0      

loaded via a namespace (and not attached):
  [1] circlize_0.4.15        plyr_1.8.8             igraph_1.4.2          
  [4] lazyeval_0.2.2         sp_1.6-0               splines_4.2.3         
  [7] listenv_0.9.0          scattermore_0.8        digest_0.6.31         
 [10] htmltools_0.5.5        fansi_1.0.4            magrittr_2.0.3        
 [13] tensor_1.5             paletteer_1.5.0        cluster_2.1.4         
 [16] ROCR_1.0-11            limma_3.54.2           remotes_2.4.2         
 [19] tzdb_0.3.0             globals_0.16.2         matrixStats_0.63.0    
 [22] R.utils_2.12.2         timechange_0.2.0       spatstat.sparse_3.0-1 
 [25] colorspace_2.1-0       ggrepel_0.9.3          xfun_0.38             
 [28] callr_3.7.3            crayon_1.5.2           jsonlite_1.8.4        
 [31] progressr_0.13.0       spatstat.data_3.0-1    survival_3.5-5        
 [34] zoo_1.8-12             glue_1.6.2             polyclip_1.10-4       
 [37] gtable_0.3.3           leiden_0.4.3           R.cache_0.16.0        
 [40] future.apply_1.10.0    shape_1.4.6            abind_1.4-5           
 [43] scales_1.2.1           spatstat.random_3.1-4  miniUI_0.1.1.1        
 [46] Rcpp_1.0.10            viridisLite_0.4.1      xtable_1.8-4          
 [49] reticulate_1.28        htmlwidgets_1.6.2      rex_1.2.1             
 [52] httr_1.4.5             RColorBrewer_1.1-3     ellipsis_0.3.2        
 [55] ica_1.0-3              farver_2.1.1           pkgconfig_2.0.3       
 [58] R.methodsS3_1.8.2      uwot_0.1.14            deldir_1.0-6          
 [61] utf8_1.2.3             janitor_2.2.0          labeling_0.4.2        
 [64] tidyselect_1.2.0       rlang_1.1.0            reshape2_1.4.4        
 [67] later_1.3.0            munsell_0.5.0          tools_4.2.3           
 [70] cli_3.6.1              ggprism_1.0.4          generics_0.1.3        
 [73] ggridges_0.5.4         evaluate_0.20          fastmap_1.1.1         
 [76] yaml_2.3.7             goftest_1.2-3          rematch2_2.1.2        
 [79] processx_3.8.1         knitr_1.42             fitdistrplus_1.1-8    
 [82] RANN_2.6.1             pbapply_1.7-0          future_1.32.0         
 [85] nlme_3.1-162           mime_0.12              ggrastr_1.0.1         
 [88] R.oo_1.25.0            xml2_1.3.3             compiler_4.2.3        
 [91] rstudioapi_0.14        beeswarm_0.4.0         plotly_4.10.1         
 [94] png_0.1-8              spatstat.utils_3.0-2   stringi_1.7.12        
 [97] cyclocomp_1.1.0        ps_1.7.5               desc_1.4.2            
[100] lattice_0.21-8         Matrix_1.5-4           vctrs_0.6.2           
[103] pillar_1.9.0           lifecycle_1.0.3        spatstat.geom_3.1-0   
[106] lmtest_0.9-40          GlobalOptions_0.1.2    RcppAnnoy_0.0.20      
[109] data.table_1.14.8      cowplot_1.1.1          irlba_2.3.5.1         
[112] httpuv_1.6.9           patchwork_1.1.2        R6_2.5.1              
[115] promises_1.2.0.1       KernSmooth_2.23-20     gridExtra_2.3         
[118] vipor_0.4.5            parallelly_1.35.0      codetools_0.2-19      
[121] MASS_7.3-58.3          rprojroot_2.0.3        withr_2.5.0           
[124] sctransform_0.4.1      parallel_4.2.3         hms_1.1.3             
[127] grid_4.2.3             rmarkdown_2.21         snakecase_0.11.0      
[130] Rtsne_0.16             spatstat.explore_3.1-0 shiny_1.7.4           
[133] ggbeeswarm_0.7.1

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 45 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "seurat_preprocessing",
  "03_3xtgad_celltype_assignment.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "seurat_preprocessing",
  "03_3xtgad_celltype_assignment.Rmd"
))
```
