---
title: "01_tf_activity"
author: "Tabea M. Soelter"
date: "2024-01-26"
output: html_document
---
**Transcription factor activity analysis using decoupleR**

__Goal__: Infer TF activity from pseudo-bulked DEGs in receiver cell types (excitatory and inhibitory neurons)

__Reproducibility__:
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.2
* HPC: Yes
    * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*Unfiltered DESeq2 outputs*
* Names:
  1. excitatory_neurons_group_6mAD_vs_6mWT_all_genes.csv
  2. excitatory_neurons_group_12mAD_vs_12mWT_all_genes.csv
  3. inhibitory_neurons_group_6mAD_vs_6mWT_all_genes.csv
  4. inhibitory_neurons_group_12mAD_vs_12mWT_all_genes.csv
* Location: results/intermediate_outputs/06_bio_activity/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * DEGs from DESeq2
* Download prior
  * CollecTRI
* Prepare for TF activity analysis
  * Create 1 matrix which has all DEGs per cell type
* Differential TF activity analysis using Multivariate Linear Model (MLM)
* Save output
  
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(decoupleR)
  library(OmnipathR)
  library(tidyverse)
  library(igraph)
  library(ComplexHeatmap)
  library(circlize)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r message=FALSE}
# DEGs in inhibitory neurons from DESeq2
inhibitory_degs_6m <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "06_bio_activity",
    "inhibitory_neurons_group_6mAD_vs_6mWT_all_genes.csv"
  )
)

inhibitory_degs_12m <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "06_bio_activity",
    "inhibitory_neurons_group_12mAD_vs_12mWT_all_genes.csv"
  )
)

# DEGs in excitatory neurons from DESeq2
excitatory_degs_6m <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "06_bio_activity",
    "excitatory_neurons_group_6mAD_vs_6mWT_all_genes.csv"
  )
)

excitatory_degs_12m <- read_csv(
  here(
    "results",
    "intermediate_outputs",
    "06_bio_activity",
    "excitatory_neurons_group_12mAD_vs_12mWT_all_genes.csv"
  )
)
```

# Download prior
* Accessed on 240208
* The connection to collectri through OmnipathR often fails
  * The next chunk loads the collectri prior I downloaded and saved
```{r}
collectri <- get_collectri(organism = "mouse", split_complexes = FALSE)
# Downloaded 38823 interactions

# Save prior
write.csv(collectri, file = here("doc", "collectri_prior_mm.csv"))
```

# Load prior
* If there are OmnipathR connection issues to collectri, load in the prior I used and saved
```{r}
collectri <- read.csv(here("doc", "collectri_prior_mm.csv"), row.names = 1)
```

# Inhibitory Neuron TF activity analysis
* We use genes that are differentially expressed (as determined by DESeq2), therefore any TF activity observed, is different between AD and WT.
  * Since WT was our base, any changes are to be interpreted with respect to AD
* We use the Multivariate Linear Model (MLM) from decoupleR
* We do this for both receiver cell types
```{r}
# Pivot DESeq2 outputs to get genes as columns with stat values
mat_in_6m <- prepare_tf_mat(inhibitory_degs_6m, timepoint = "6m")

mat_in_12m <- prepare_tf_mat(inhibitory_degs_12m, timepoint = "12m")

# Combine time points
df_in <- merge(mat_in_6m, mat_in_12m, all = TRUE)

# replace NAs with 0
df_in[is.na(df_in)] <- 0

# turn into matrix
mat_in <- as.matrix(df_in)

rownames(mat_in) <- c("6m", "12m")

# TF activity analysis
tf_acts_in <- run_mlm(
  mat = t(mat_in),
  network = collectri
)
```

# Excitatory Neuron TF activity analysis
```{r}
# Pivot DESeq2 outputs to get genes as columns with stat values
mat_ex_6m <- prepare_tf_mat(excitatory_degs_6m, timepoint = "6m")

mat_ex_12m <- prepare_tf_mat(excitatory_degs_12m, timepoint = "12m")

# Combine time points
df_ex <- merge(mat_ex_6m, mat_ex_12m, all = TRUE)

# replace NAs with 0
df_ex[is.na(df_ex)] <- 0

# turn into matrix
mat_ex <- as.matrix(df_ex)

rownames(mat_ex) <- c("6m", "12m")

# TF activity analysis
tf_acts_ex <- run_mlm(
  mat = t(mat_ex),
  network = collectri
)
```

# Save TF activity outputs
```{r}
saveRDS(
  tf_acts_in,
  here(
    "results",
    "final_outputs",
    "05_bio_activity",
    "inhibitory_tf_activity.rds"
  )
)

saveRDS(
  tf_acts_ex,
  here(
    "results",
    "final_outputs",
    "05_bio_activity",
    "excitatory_tf_activity.rds"
  )
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2           styler_1.9.1          here_1.0.1           
 [4] circlize_0.4.15       ComplexHeatmap_2.14.0 igraph_1.4.2         
 [7] lubridate_1.9.2       forcats_1.0.0         stringr_1.5.0        
[10] dplyr_1.1.1           purrr_1.0.1           readr_2.1.4          
[13] tidyr_1.3.0           tibble_3.2.1          ggplot2_3.4.2        
[16] tidyverse_2.0.0       OmnipathR_3.9.8       decoupleR_2.9.1      

loaded via a namespace (and not attached):
 [1] matrixStats_0.63.0  bit64_4.0.5         doParallel_1.0.17   RColorBrewer_1.1-3 
 [5] progress_1.2.2      httr_1.4.5          rprojroot_2.0.3     R.cache_0.16.0     
 [9] tools_4.2.3         backports_1.4.1     utf8_1.2.3          R6_2.5.1           
[13] lazyeval_0.2.2      BiocGenerics_0.44.0 colorspace_2.1-0    GetoptLong_1.0.5   
[17] withr_2.5.0         processx_3.8.1      tidyselect_1.2.0    prettyunits_1.1.1  
[21] bit_4.0.5           curl_5.0.0          compiler_4.2.3      cli_3.6.1          
[25] rvest_1.0.3         xml2_1.3.3          desc_1.4.2          scales_1.2.1       
[29] checkmate_2.1.0     callr_3.7.3         rappdirs_0.3.3      digest_0.6.31      
[33] R.utils_2.12.2      rmarkdown_2.21      pkgconfig_2.0.3     htmltools_0.5.5    
[37] parallelly_1.35.0   fastmap_1.1.1       rlang_1.1.0         GlobalOptions_0.1.2
[41] readxl_1.4.2        rstudioapi_0.14     shape_1.4.6         generics_0.1.3     
[45] jsonlite_1.8.4      vroom_1.6.1         R.oo_1.25.0         magrittr_2.0.3     
[49] Matrix_1.5-4        Rcpp_1.0.10         munsell_0.5.0       S4Vectors_0.36.2   
[53] fansi_1.0.4         logger_0.2.2        lifecycle_1.0.3     R.methodsS3_1.8.2  
[57] stringi_1.7.12      yaml_2.3.7          parallel_4.2.3      crayon_1.5.2       
[61] lattice_0.21-8      hms_1.1.3           ps_1.7.5            knitr_1.42         
[65] pillar_1.9.0        rjson_0.2.21        codetools_0.2-19    stats4_4.2.3       
[69] glue_1.6.2          evaluate_0.20       rex_1.2.1           remotes_2.4.2      
[73] png_0.1-8           vctrs_0.6.2         tzdb_0.3.0          foreach_1.5.2      
[77] cellranger_1.1.0    gtable_0.3.3        clue_0.3-64         xfun_0.38          
[81] cyclocomp_1.1.0     later_1.3.0         iterators_1.0.14    IRanges_2.32.0     
[85] cluster_2.1.4       timechange_0.2.0

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "bio_activity",
  "01_tf_activity.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "bio_activity",
  "01_tf_activity.Rmd"
))
```
