---
title: "01_ambient_RNA_removal"
author: "Tabea M. Soelter"
date: "2023-10-23"
output: html_document
---
**Ambient RNA removal**

__Goal__: Remove ambient RNA for downstream analyses to guarantee highest quality of the dataset.
  
__Reproducibility__: 
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.0
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 85GB per CPU

__Data__: 
*Cell Ranger outputs*
* Name: N/A
* Location: data/CellRangerCounts/pre_soupX/
 
__Analysis Plan__: 
* Load necessary packages 
* Load h5 data from Cell Ranger
* Make Soup Channel object
* Create Seurat object
* Cluster Seurat object
* Profile the soup
* Create integer matrix
* Save filtered objects

__Analysis__:
# load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(Seurat)
  library(SoupX)
  library(DropletUtils)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Ambient RNA removal
* Here I am using a function which takes in the h5 files from Cell Ranger and performs all necessary steps to remove ambient RNA. Ultimately, filtered files necessary for downstream pre-processing are saved by sample.
```{r}
remove_ambientRNA(
  inputs = here(
    "data",
    "CellRangerCounts",
    "pre_soupX/"
  ),
  outputs = here(
    "data",
    "CellRangerCounts",
    "post_soupX/"
  ),
  plots = here(
    "results",
    "intermediate_outputs",
    "01_soupX/"
  )
)
```

# Session Info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats4    stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2                 styler_1.9.1               
 [3] here_1.0.1                  DropletUtils_1.18.1        
 [5] SingleCellExperiment_1.20.1 SummarizedExperiment_1.28.0
 [7] Biobase_2.58.0              GenomicRanges_1.50.2       
 [9] GenomeInfoDb_1.34.9         IRanges_2.32.0             
[11] S4Vectors_0.36.2            BiocGenerics_0.44.0        
[13] MatrixGenerics_1.10.0       matrixStats_0.63.0         
[15] SoupX_1.6.2                 SeuratObject_4.1.3         
[17] Seurat_4.3.0               

loaded via a namespace (and not attached):
  [1] plyr_1.8.8                igraph_1.4.2              lazyeval_0.2.2           
  [4] sp_1.6-0                  splines_4.2.3             BiocParallel_1.32.6      
  [7] listenv_0.9.0             scattermore_0.8           ggplot2_3.4.2            
 [10] digest_0.6.31             htmltools_0.5.5           fansi_1.0.4              
 [13] magrittr_2.0.3            tensor_1.5                cluster_2.1.4            
 [16] ROCR_1.0-11               remotes_2.4.2             limma_3.54.2             
 [19] globals_0.16.2            R.utils_2.12.2            spatstat.sparse_3.0-1    
 [22] colorspace_2.1-0          ggrepel_0.9.3             xfun_0.38                
 [25] dplyr_1.1.1               crayon_1.5.2              callr_3.7.3              
 [28] RCurl_1.98-1.12           jsonlite_1.8.4            progressr_0.13.0         
 [31] spatstat.data_3.0-1       survival_3.5-5            zoo_1.8-12               
 [34] glue_1.6.2                polyclip_1.10-4           gtable_0.3.3             
 [37] zlibbioc_1.44.0           XVector_0.38.0            leiden_0.4.3             
 [40] DelayedArray_0.24.0       R.cache_0.16.0            Rhdf5lib_1.20.0          
 [43] future.apply_1.10.0       HDF5Array_1.26.0          abind_1.4-5              
 [46] scales_1.2.1              edgeR_3.40.2              spatstat.random_3.1-4    
 [49] miniUI_0.1.1.1            Rcpp_1.0.10               viridisLite_0.4.1        
 [52] xtable_1.8-4              dqrng_0.3.0               reticulate_1.28          
 [55] bit_4.0.5                 rex_1.2.1                 htmlwidgets_1.6.2        
 [58] httr_1.4.5                RColorBrewer_1.1-3        ellipsis_0.3.2           
 [61] ica_1.0-3                 scuttle_1.8.4             pkgconfig_2.0.3          
 [64] R.methodsS3_1.8.2         uwot_0.1.14               deldir_1.0-6             
 [67] locfit_1.5-9.7            utf8_1.2.3                tidyselect_1.2.0         
 [70] rlang_1.1.0               reshape2_1.4.4            later_1.3.0              
 [73] munsell_0.5.0             tools_4.2.3               cli_3.6.1                
 [76] generics_0.1.3            ggridges_0.5.4            evaluate_0.20            
 [79] stringr_1.5.0             fastmap_1.1.1             yaml_2.3.7               
 [82] goftest_1.2-3             bit64_4.0.5               processx_3.8.1           
 [85] knitr_1.42                fitdistrplus_1.1-8        purrr_1.0.1              
 [88] RANN_2.6.1                pbapply_1.7-0             future_1.32.0            
 [91] nlme_3.1-162              sparseMatrixStats_1.10.0  mime_0.12                
 [94] R.oo_1.25.0               xml2_1.3.3                hdf5r_1.3.8              
 [97] compiler_4.2.3            rstudioapi_0.14           plotly_4.10.1            
[100] png_0.1-8                 spatstat.utils_3.0-2      tibble_3.2.1             
[103] stringi_1.7.12            ps_1.7.5                  cyclocomp_1.1.0          
[106] desc_1.4.2                lattice_0.21-8            Matrix_1.5-4             
[109] vctrs_0.6.2               pillar_1.9.0              lifecycle_1.0.3          
[112] rhdf5filters_1.10.1       spatstat.geom_3.1-0       lmtest_0.9-40            
[115] RcppAnnoy_0.0.20          data.table_1.14.8         cowplot_1.1.1            
[118] bitops_1.0-7              irlba_2.3.5.1             httpuv_1.6.9             
[121] patchwork_1.1.2           R6_2.5.1                  promises_1.2.0.1         
[124] KernSmooth_2.23-20        gridExtra_2.3             parallelly_1.35.0        
[127] codetools_0.2-19          MASS_7.3-58.3             rhdf5_2.42.1             
[130] rprojroot_2.0.3           withr_2.5.0               sctransform_0.4.1        
[133] GenomeInfoDbData_1.2.9    parallel_4.2.3            grid_4.2.3               
[136] beachmat_2.14.2           tidyr_1.3.0               rmarkdown_2.21           
[139] DelayedMatrixStats_1.20.0 Rtsne_0.16                spatstat.explore_3.1-0   
[142] shiny_1.7.4 

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 45 mins
# Hands-on time however is 5 minutes!
```

# Reproducibility
```{r eval=FALSE, include=FALSE}
# styler
style_file(here(
  "src",
  "soupX",
  "01_ambient_RNA_removal.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "soupX",
  "01_ambient_RNA_removal.Rmd"
))
```
