---
title: "01_differential_ccc"
author: "Tabea M. Soelter"
date: "2023-11-13"
output: html_document
---
**Differential Cell-Cell Communication Analysis using MultiNicheNet**

__Goal__: Perform MultiNicheNet analyses between AD and WT to identify ligands, receptors, and potential downstream target genes between glia (Astrocytes, Oligodendrocytes, Microglia, and OPCs - senders) and Excitatory/Inhibitory Neurons (receivers).
  
__Reproducibility__: 
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.1
* HPC: Yes
  * Resources: long partition (150hrs), 1 GPU, 14 CPUs, 45GB per CPU

__Data__:
*Seurat object*
* Name: processed_seurat.rds
* Location: data/seurat/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * Seurat object
* Load and filter NicheNet v2 prior
* Prepare for NicheNet
  * Determine cut-offs and comparisons
* Differential NicheNet using MultiNicheNet
  * Wrapper function
* Save output

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(nichenetr)
  library(multinichenetr)
  library(SingleCellExperiment)
  library(tidyverse)
  library(tibble)
  library(Seurat)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
processed_seurat <- readRDS(here(
  "data",
  "seurat",
  "processed_seurat.rds"
))
```

# NicheNet-v2 prior
* downloaded here on 231115: https://zenodo.org/record/7074291
```{r}
lr_network <- readRDS(here(
  "data",
  "ccc",
  "nichenet_v2_prior",
  "lr_network_mouse_21122021.rds"
)) %>%
  dplyr::rename(ligand = from, receptor = to) %>%
  distinct(ligand, receptor) %>%
  mutate(ligand = make.names(ligand), receptor = make.names(receptor))

ligand_target_matrix <- readRDS(here(
  "data",
  "ccc",
  "nichenet_v2_prior",
  "ligand_target_matrix_nsga2r_final_mouse.rds"
))

# format row and column names
colnames(ligand_target_matrix) <- colnames(ligand_target_matrix) %>%
  make.names()
rownames(ligand_target_matrix) <- rownames(ligand_target_matrix) %>%
  make.names()
```

# Convert to SCE object
```{r}
sce <- Seurat::as.SingleCellExperiment(processed_seurat, assay = "RNA")
sce <- alias_to_symbol_SCE(sce, "mouse") %>% makenames_SCE()
```

# Create group column
As I want to compare CCC between AD and WT for multiple timepoints, I will create a group column which combines the following information:
  * orig.ident = condition (AD/WT)
  * timepoint = time point (6m/12m)
```{r}
colData(sce)$group <- factor(
  paste0(colData(sce)$timepoint, colData(sce)$orig.ident)
)
```

# Make names syntactically valid
```{r}
sce <- make_names_valid(sce)
```

# Set variables for wrapper
```{r}
# sample IDs
sample_id <- "sample"

# condition information
group_id <- "group"

# cell type information
celltype_id <- "ident"

# Other DE variables were denoted as `NA`
covariates <- NA
batches <- NA

# Since we have small sample numbers, we do not use adjusted p-values
empirical_pval <- FALSE
p_val_adj <- FALSE

# Number of cores of the HPC session
cores_system <- 14
```

# Prioritization weights
* These weights are recommended by the developers of MultiNicheNet. There are more detailed descriptions in the following step-by-step vignette:
  * https://github.com/saeyslab/multinichenetr/blob/main/vignettes/pairwise_analysis_MISC.md 
```{r}
prioritizing_weights_DE <- c(
  "de_ligand" = 1,
  "de_receptor" = 1
)

prioritizing_weights_activity <- c(
  "activity_scaled" = 2
)

prioritizing_weights_expression_specificity <- c(
  "exprs_ligand" = 2,
  "exprs_receptor" = 2
)

prioritizing_weights_expression_sufficiency <- c(
  "frac_exprs_ligand_receptor" = 1
)

prioritizing_weights_relative_abundance <- c(
  "abund_sender" = 0,
  "abund_receiver" = 0
)

# Combine all weights into 1 variable
prioritizing_weights <- c(
  prioritizing_weights_DE,
  prioritizing_weights_activity,
  prioritizing_weights_expression_specificity,
  prioritizing_weights_expression_sufficiency,
  prioritizing_weights_relative_abundance
)
```

# Contrast table
* The format of the contrast_oi variable requires no spaces, therefore ensure that you don't add spaces to contrast_oi.
```{r}
contrasts_oi <- c("'X6mAD-X6mWT','X6mWT-X6mAD','X12mAD-X12mWT','X12mWT-X12mAD'")

contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)
```

# CCC inference
```{r}
# path for abundance expression plots
results_path <- "results/intermediate_outputs/04_ccc"

# this runs for approx 50 minutes
multinichenet_output <- multinichenet_wrapper(
  object = sce,
  results_path = results_path,
  celltype_id = celltype_id,
  sample_id = sample_id,
  group_id = group_id,
  lr_network = lr_network,
  batches = batches,
  contrasts_oi = contrasts_oi,
  contrast_tbl = contrast_tbl,
  covariates = covariates,
  empirical_pval = empirical_pval,
  p_val_adj = p_val_adj,
  cores_system = cores_system,
  ligand_target_matrix = ligand_target_matrix,
  prioritizing_weights = prioritizing_weights
)
```

# Save outputs
```{r}
saveRDS(multinichenet_output, here(
  "data",
  "ccc",
  "multinichenet_output.rds"
))
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats4    stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2                 styler_1.9.1               
 [3] here_1.0.1                  SeuratObject_4.1.3         
 [5] Seurat_4.3.0                lubridate_1.9.2            
 [7] forcats_1.0.0               stringr_1.5.0              
 [9] dplyr_1.1.1                 purrr_1.0.1                
[11] readr_2.1.4                 tidyr_1.3.0                
[13] tibble_3.2.1                ggplot2_3.4.2              
[15] tidyverse_2.0.0             SingleCellExperiment_1.20.1
[17] SummarizedExperiment_1.28.0 Biobase_2.58.0             
[19] GenomicRanges_1.50.2        GenomeInfoDb_1.34.9        
[21] IRanges_2.32.0              S4Vectors_0.36.2           
[23] BiocGenerics_0.44.0         MatrixGenerics_1.10.0      
[25] matrixStats_0.63.0          multinichenetr_1.0.3       
[27] nichenetr_2.0.4            

loaded via a namespace (and not attached):
  [1] rsvd_1.0.5                Hmisc_5.0-1               ica_1.0-3                
  [4] ps_1.7.5                  class_7.3-21              rprojroot_2.0.3          
  [7] foreach_1.5.2             lmtest_0.9-40             crayon_1.5.2             
 [10] rbibutils_2.2.13          MASS_7.3-58.3             nlme_3.1-162             
 [13] backports_1.4.1           sva_3.46.0                rlang_1.1.0              
 [16] XVector_0.38.0            caret_6.0-94              ROCR_1.0-11              
 [19] irlba_2.3.5.1             callr_3.7.3               nloptr_2.0.3             
 [22] limma_3.54.2              scater_1.26.1             BiocParallel_1.32.6      
 [25] rjson_0.2.21              bit64_4.0.5               glue_1.6.2               
 [28] sctransform_0.4.1         processx_3.8.1            pbkrtest_0.5.2           
 [31] parallel_4.2.3            vipor_0.4.5               spatstat.sparse_3.0-1    
 [34] AnnotationDbi_1.60.2      UpSetR_1.4.0              muscat_1.12.1            
 [37] spatstat.geom_3.1-0       tidyselect_1.2.0          fitdistrplus_1.1-8       
 [40] variancePartition_1.28.9  XML_3.99-0.14             zoo_1.8-12               
 [43] ggpubr_0.6.0              xtable_1.8-4              magrittr_2.0.3           
 [46] evaluate_0.20             Rdpack_2.4                scuttle_1.8.4            
 [49] cli_3.6.1                 zlibbioc_1.44.0           rstudioapi_0.14          
 [52] miniUI_0.1.1.1            sp_1.6-0                  rpart_4.1.19             
 [55] aod_1.3.2                 locfdr_1.1-8              shiny_1.7.4              
 [58] BiocSingular_1.14.0       xfun_0.38                 clue_0.3-64              
 [61] cluster_2.1.4             caTools_1.18.2            tidygraph_1.2.3          
 [64] KEGGREST_1.38.0           clusterGeneration_1.3.7   ggrepel_0.9.3            
 [67] listenv_0.9.0             Biostrings_2.66.0         png_0.1-8                
 [70] future_1.32.0             ipred_0.9-14              withr_2.5.0              
 [73] bitops_1.0-7              ggforce_0.4.1             plyr_1.8.8               
 [76] hardhat_1.3.0             dqrng_0.3.0               e1071_1.7-13             
 [79] pROC_1.18.0               coda_0.19-4               pillar_1.9.0             
 [82] gplots_3.1.3              GlobalOptions_0.1.2       cachem_1.0.7             
 [85] GetoptLong_1.0.5          RUnit_0.4.32              DelayedMatrixStats_1.20.0
 [88] vctrs_0.6.2               ellipsis_0.3.2            generics_0.1.3           
 [91] lava_1.7.2.1              tools_4.2.3               foreign_0.8-84           
 [94] remaCor_0.0.11            beeswarm_0.4.0            munsell_0.5.0            
 [97] tweenr_2.0.2              emmeans_1.8.5             proxy_0.4-27             
[100] DelayedArray_0.24.0       fastmap_1.1.1             compiler_4.2.3           
[103] abind_1.4-5               httpuv_1.6.9              plotly_4.10.1            
[106] GenomeInfoDbData_1.2.9    prodlim_2023.03.31        gridExtra_2.3            
[109] glmmTMB_1.1.7             edgeR_3.40.2              lattice_0.21-8           
[112] ggnewscale_0.4.8          deldir_1.0-6              visNetwork_2.1.2         
[115] utf8_1.2.3                later_1.3.0               recipes_1.0.5            
[118] jsonlite_1.8.4            scales_1.2.1              ScaledMatrix_1.6.0       
[121] carData_3.0-5             pbapply_1.7-0             sparseMatrixStats_1.10.0 
[124] estimability_1.4.1        genefilter_1.80.3         lazyeval_0.2.2           
[127] promises_1.2.0.1          car_3.1-2                 doParallel_1.0.17        
[130] R.utils_2.12.2            goftest_1.2-3             spatstat.utils_3.0-2     
[133] reticulate_1.28           checkmate_2.1.0           cyclocomp_1.1.0          
[136] rmarkdown_2.21            cowplot_1.1.1             blme_1.0-5               
[139] statmod_1.5.0             Rtsne_0.16                uwot_0.1.14              
[142] igraph_1.4.2              survival_3.5-5            numDeriv_2016.8-1.1      
[145] yaml_2.3.7                htmltools_0.5.5           memoise_2.0.1            
[148] locfit_1.5-9.7            graphlayouts_0.8.4        viridisLite_0.4.1        
[151] digest_0.6.31             RhpcBLASctl_0.23-42       mime_0.12                
[154] RSQLite_2.3.1             future.apply_1.10.0       remotes_2.4.2            
[157] data.table_1.14.8         blob_1.2.4                R.oo_1.25.0              
[160] DiagrammeR_1.0.9          labeling_0.4.2            splines_4.2.3            
[163] Formula_1.2-5             RCurl_1.98-1.12           broom_1.0.4              
[166] hms_1.1.3                 colorspace_2.1-0          base64enc_0.1-3          
[169] ggbeeswarm_0.7.1          shape_1.4.6               nnet_7.3-18              
[172] Rcpp_1.0.10               RANN_2.6.1                mvtnorm_1.1-3            
[175] circlize_0.4.15           fansi_1.0.4               tzdb_0.3.0               
[178] parallelly_1.35.0         ModelMetrics_1.2.2.2      R6_2.5.1                 
[181] grid_4.2.3                factoextra_1.0.7          ggridges_0.5.4           
[184] lifecycle_1.0.3           bluster_1.8.0             ggsignif_0.6.4           
[187] minqa_1.2.5               leiden_0.4.3              Matrix_1.5-4             
[190] desc_1.4.2                RcppAnnoy_0.0.20          RColorBrewer_1.1-3       
[193] iterators_1.0.14          spatstat.explore_3.1-0    TMB_1.9.4                
[196] gower_1.0.1               R.cache_0.16.0            htmlwidgets_1.6.2        
[199] beachmat_2.14.2           polyclip_1.10-4           shadowtext_0.1.2         
[202] timechange_0.2.0          mgcv_1.8-42               ComplexHeatmap_2.14.0    
[205] globals_0.16.2            htmlTable_2.4.1           patchwork_1.1.2          
[208] spatstat.random_3.1-4     progressr_0.13.0          codetools_0.2-19         
[211] metapod_1.6.0             randomForest_4.7-1.1      gtools_3.9.4             
[214] prettyunits_1.1.1         R.methodsS3_1.8.2         gtable_0.3.3             
[217] DBI_1.1.3                 tensor_1.5                httr_1.4.5               
[220] KernSmooth_2.23-20        stringi_1.7.12            progress_1.2.2           
[223] reshape2_1.4.4            farver_2.1.1              annotate_1.76.0          
[226] viridis_0.6.2             rex_1.2.1                 fdrtool_1.2.17           
[229] timeDate_4022.108         xml2_1.3.3                boot_1.3-28.1            
[232] BiocNeighbors_1.16.0      lme4_1.1-32               geneplotter_1.76.0       
[235] scattermore_0.8           scran_1.26.2              DESeq2_1.38.3            
[238] bit_4.0.5                 spatstat.data_3.0-1       ggraph_2.1.0             
[241] pkgconfig_2.0.3           lmerTest_3.1-3            rstatix_0.7.2            
[244] knitr_1.42

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 1 hour
# Hands-on time however is between 10-15 minutes!
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "ccc",
  "01_differential_ccc.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "ccc",
  "01_differential_ccc.Rmd"
))
```
