---
title: "02_pathway_analysis_targets"
author: "Tabea M. Soelter"
date: "2024-01-08"
output: html_document
---
**Pathway Analysis of predicted target genes using gprofiler2**

__Goal__: Perform pathway analyses for genes that were predicted as potential downstream targets in inhibitory and excitatory neurons of CCC interactions altered in AD. 
  
__Reproducibility__: 
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.1
* HPC: Yes
  * Resources: long partition (150hrs), 1 GPU, 6 CPUs, 65GB per CPU

__Data__:
*Seurat object*
* Name: multinichenet_output.rds
* Location: data/ccc/
 
__Analysis Plan__:
* Load necessary packages 
* Load data
  * MultiNicheNet object
* Filter object to get target genes
* Pathway Analysis using gprofiler2
  * Plotting of pathways
  * Saving plots
* Save FEA outputs

# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(gprofiler2)
  library(tidyverse)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
output <- readRDS(here("data", "ccc", "multinichenet_output.rds"))
```

# Set filtering criteria
* These are continuously the same throughout the project
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# set groups to filter for (in this case all of them)
group_oi <- c("X6mWT", "X6mAD", "X12mWT", "X12mAD")
```

# Filter MultiNicheNet output
```{r}
prior <- filter_nichenet(output)

# select needed columns
prior_filt <- prior %>%
  select("id_target", "sender", "target", "group")
```

# Compile background genes
* We are including our custom background gene set for our FEA analysis:
  * Included are all genes expressed in receivers that could have been potential target genes
  * This include genes that are not significantly differentially expressed and genes in our query
```{r}
# access list of all possible target genes in MultiNicheNet output
all_genes_df <- output$celltype_de

# grab list of targets
all_genes <- all_genes_df %>%
  filter(cluster_id %in% c("Excitatory.Neurons", "Inhibitory.Neurons")) %>%
  select(gene) %>%
  unique() %>%
  pull(gene)
```

# FEA: Shared genes
* Shared = genes shared between the 6- and 12-month time points within the same condition
* The fea function is a wrapper for code adapted from Elizabeth J. Wilk
```{r}
# Filter duplicate interactions
shared <- prior_filt %>%
  group_by(id_target) %>%
  filter(n() > 1)

# New column based on group column (i.e. Both 6mAD & 12mAD = AD, same for WT)
shared$condition <- substring(
  shared$group,
  nchar(shared$group) - 1,
  nchar(shared$group)
)

# Remove old column to get unique interactions
shared <- shared %>%
  select(-group)

shared <- unique(shared)

# Filter and select columns of interest
shared_filt <- shared %>%
  ungroup() %>%
  select(target, condition) %>%
  unique()

# Filter for AD-specific genes and perform FEA
shared_ad <- shared_filt %>%
  filter(condition == "AD") %>%
  select(target) %>%
  as.list()

fea_res_ad <- fea(shared_ad$target,
  bg_genes = all_genes
) %>%
  mutate(condition = "AD")

# Filter for WT-specific genes and perform FEA
shared_wt <- shared_filt %>%
  filter(condition == "WT") %>%
  select(target) %>%
  as.list()

fea_res_wt <- fea(shared_wt$target,
  bg_genes = all_genes
) %>%
  mutate(condition = "WT")

# combine for plotting purposes
fea_res_shared <- rbind(fea_res_ad, fea_res_wt) # no terms for WT
```

# Plotting shared pathways
* This only includes pathways in AD, as WT had no terms returned to its query
  * The query was very small, so this is not surprising
```{r}
barplot_shared <- ggplot(
  fea_res_shared,
  aes(
    y = recall,
    x = reorder(term_name, p_value),
    fill = p_value
  )
) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_distiller(palette = "Greens") +
  labs(x = "Functional Enrichment Terms", y = "Recall", fill = "adj p-value") +
  ggtitle("Top Enriched Terms for Shared Target Genes") +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(face = "bold", color = "black", size = 10),
    text = element_text(face = "bold"),
    axis.title = element_text(face = "bold", size = 12)
  )

barplot_shared

ggsave(
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "fea_shared_ad_barplot.png"
  ),
  barplot_shared,
  dpi = 400,
  width = 10,
  height = 5,
  bg = "#ffffff"
)
```

# Filter for unique genes 
* shared_filt includes all targets that are somehow shared 
  * This includes interactions shared between 6m and 12m
  * I am filtering condition and time point pairings to exclude any of those above targets (even if the LR pair is different than in the shared interactions, bc the target is still shared, implying the same downstream effect) 
```{r}
targets_6m_ad <- prior_filt %>%
  filter(group == "X6mAD") %>%
  filter(!target %in% shared_filt$target) # 75

targets_6m_wt <- prior_filt %>%
  filter(group == "X6mWT") %>%
  filter(!target %in% shared_filt$target) # 66

targets_12m_ad <- prior_filt %>%
  filter(group == "X12mAD") %>%
  filter(!target %in% shared_filt$target) # 1624

targets_12m_wt <- prior_filt %>%
  filter(group == "X12mWT") %>%
  filter(!target %in% shared_filt$target) # 429
```

# FEA: Unique genes - 6-month
```{r}
# FEA on 6m unique genes by condition, then combine for plotting
fea_res_6m_ad_unique <- fea(targets_6m_ad$target,
  bg_genes = all_genes
) %>%
  mutate(condition = "AD")

fea_res_6m_wt_unique <- fea(targets_6m_wt$target,
  bg_genes = all_genes
) %>%
  mutate(condition = "WT")

fea_res_6m <- rbind(fea_res_6m_ad_unique, fea_res_6m_wt_unique)
```

# Plotting 6-month pathways
```{r}
bubbleplot_6m_unique <- ggplot(
  fea_res_6m,
  aes(
    x = condition,
    y = reorder(term_name, p_value),
    size = recall,
    fill = p_value
  )
) +
  geom_point(alpha = 0.7, shape = 21) +
  scale_size(range = c(2, 10), name = "Recall") +
  scale_fill_distiller(palette = "Greens") +
  labs(x = "Condition", y = "Functional Enrichment Terms") +
  theme_minimal() +
  ggtitle("Top Enriched Terms for Target\nGenes at 6-month time point") +
  theme_minimal() +
  theme(
    axis.text = element_text(face = "bold"),
    text = element_text(face = "bold")
  )

bubbleplot_6m_unique

ggsave(
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "fea_6m_bubbleplot.png"
  ),
  bubbleplot_6m_unique,
  dpi = 400,
  width = 8,
  height = 7,
  bg = "#ffffff"
)
```

# FEA: Unique genes - 12-month
```{r}
# FEA on 12m unique genes by condition, then combine for plotting
fea_res_12m_ad_unique <- fea(targets_12m_ad$target,
  bg_genes = all_genes
) %>%
  mutate(condition = "AD")

fea_res_12m_wt_unique <- fea(targets_12m_wt$target,
  bg_genes = all_genes
) %>%
  mutate(condition = "WT")

fea_res_12m <- rbind(fea_res_12m_ad_unique, fea_res_12m_wt_unique)
```

# Plotting 12-month pathways
```{r}
bubbleplot_12m_unique <- ggplot(
  fea_res_12m,
  aes(
    x = condition,
    y = reorder(term_name, p_value),
    size = recall,
    fill = p_value
  )
) +
  geom_point(alpha = 0.7, shape = 21) +
  scale_size(range = c(2, 10), name = "Recall") +
  scale_fill_distiller(palette = "Greens") +
  labs(x = "Condition", y = "Functional Enrichment Terms") +
  theme_minimal() +
  ggtitle("Top Enriched Terms for Target\nGenes at 12-month time point") +
  theme_minimal() +
  theme(
    axis.text = element_text(face = "bold"),
    text = element_text(face = "bold")
  )

bubbleplot_12m_unique

ggsave(
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "fea_12m_bubbleplot.png"
  ),
  bubbleplot_12m_unique,
  dpi = 400,
  width = 13,
  height = 7,
  bg = "#ffffff"
)
```

# Save filtered FEA outputs
```{r}
saveRDS(fea_res_shared, file = here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "fea_res_shared.rds"
))

saveRDS(fea_res_6m, file = here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "fea_res_6m.rds"
))

saveRDS(fea_res_12m, file = here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "fea_res_12m.rds"
))
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2      styler_1.9.1     here_1.0.1       lubridate_1.9.2 
 [5] forcats_1.0.0    stringr_1.5.0    dplyr_1.1.1      purrr_1.0.1     
 [9] readr_2.1.4      tidyr_1.3.0      tibble_3.2.1     ggplot2_3.4.2   
[13] tidyverse_2.0.0  gprofiler2_0.2.1

loaded via a namespace (and not attached):
 [1] ps_1.7.5           rprojroot_2.0.3    digest_0.6.31      utf8_1.2.3        
 [5] R6_2.5.1           evaluate_0.20      httr_1.4.5         pillar_1.9.0      
 [9] rlang_1.1.0        lazyeval_0.2.2     rstudioapi_0.14    data.table_1.14.8 
[13] callr_3.7.3        R.utils_2.12.2     R.oo_1.25.0        rmarkdown_2.21    
[17] textshaping_0.3.6  labeling_0.4.2     desc_1.4.2         htmlwidgets_1.6.2 
[21] RCurl_1.98-1.12    munsell_0.5.0      compiler_4.2.3     xfun_0.38         
[25] systemfonts_1.0.4  pkgconfig_2.0.3    htmltools_0.5.5    tidyselect_1.2.0  
[29] fansi_1.0.4        viridisLite_0.4.1  crayon_1.5.2       tzdb_0.3.0        
[33] withr_2.5.0        bitops_1.0-7       R.methodsS3_1.8.2  grid_4.2.3        
[37] jsonlite_1.8.4     gtable_0.3.3       lifecycle_1.0.3    magrittr_2.0.3    
[41] scales_1.2.1       cli_3.6.1          stringi_1.7.12     farver_2.1.1      
[45] remotes_2.4.2      rex_1.2.1          xml2_1.3.3         ragg_1.2.5        
[49] generics_0.1.3     vctrs_0.6.2        cyclocomp_1.1.0    RColorBrewer_1.1-3
[53] tools_4.2.3        R.cache_0.16.0     glue_1.6.2         hms_1.1.3         
[57] processx_3.8.1     fastmap_1.1.1      yaml_2.3.7         timechange_0.2.0  
[61] colorspace_2.1-0   plotly_4.10.1      knitr_1.42 

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: 10 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "ccc",
  "02_pathway_analysis_targets.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "ccc",
  "02_pathway_analysis_targets.Rmd"
))
```
