---
title: "03_jaccard_similarity"
author: "Tabea M. Soelter"
date: "2024-01-09"
output: html_document
---
**Jaccard Similarity Index (JI) for CCC genes across cell types and time points**

__Goal__: Calculate similarity, using the Jaccard Similarity Index, of identified ligands, receptors, and target genes across cell types and time points. This allows us to identify cell-type-specific and cell-type-agnostic CCC mechanisms across time points.
  
__Reproducibility__: 
* GitHub: lasseignelab/230418_TS_AgingCCC
* Docker: tsoelter/rstudio_aging_ccc
    * Version: 1.0.1
* HPC: Yes
  * Resources: long partition (150hrs), 6 CPUs, 65GB per CPU

__Data__:
* Name: multinichenet_output.rds
* Location: data/ccc/
 
__Analysis Plan__:
* Load necessary packages 
* Load data 
* Prepare input for Jaccard Index calculation
  * Create dfs with 2 columns: gene, cell type
    * gene refers to either the ligands, receptors, or target genes
* Calculate Jaccard Index
  * Do this for both Ex and In Neurons

__Analysis__:
# Load packages
```{r}
# Time tracking
ptm <- proc.time()

# Setting seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(styler)
  library(lintr)
})

source(here("src", "functions_CCCin3xTgAD.R"))
```

# Load data
```{r}
output <- readRDS(here("data", "ccc", "multinichenet_output.rds"))
```

# Set Filtering Criteria
* These are the same as previously:
  * Senders and receivers of interest
  * Top number of predicted targets for predicted interactions
  * Groups of interest (condition/time point combinations)
```{r}
# number of targets
top_n_target <- 250

# contrast table
contrast_tbl <- tibble(
  contrast = c("X6mAD-X6mWT", "X6mWT-X6mAD", "X12mAD-X12mWT", "X12mWT-X12mAD"),
  group = c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
)

# set receiver and sender cell types of interest
receiver_oi <- c("Excitatory.Neurons", "Inhibitory.Neurons")
sender_oi <- c("Astrocytes", "Microglia", "OPCs", "Oligodendrocytes")

# set groups
group_oi <- c("X6mAD", "X6mWT", "X12mAD", "X12mWT")
```

# Filter MultiNicheNet output
* Filter by ligand activity and regulatory potential (Pearson and Spearman correlation coefficients)
```{r warning=FALSE}
# filter output
lrt_cor_filtered <- filter_nichenet(output)

# Rename receiver cell types
lrt_cor_filtered$receiver <- ifelse(
  lrt_cor_filtered$receiver == "Inhibitory.Neurons",
  "Inhibitory Neurons",
  "Excitatory Neurons"
)

# Check receivers were renamed and number
table(lrt_cor_filtered$receiver) # Ex 607 In 1995
```

# Calculate JI (Between senders)
```{r}
# set receivers again, since I modified them
receivers_oi <- c("Excitatory Neurons", "Inhibitory Neurons")

# ligands
jaccard_results_ligands <- calculate_jaccard_senders(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "ligands"
)

# receptors
jaccard_results_receptors <- calculate_jaccard_senders(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "receptors"
)

# targets
jaccard_results_targets <- calculate_jaccard_senders(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "targets"
)
```

# Calculate JI (Between receivers)
```{r warning=FALSE}
jaccard_receptors_receiver <- calculate_jaccard_receivers(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "receptor"
)

jaccard_targets_receiver <- calculate_jaccard_receivers(lrt_cor_filtered,
  senders = sender_oi,
  receivers = receivers_oi,
  type = "target"
)
```

# Arrange JI outputs for plotting
```{r}
# receptors
jaccard_receptors <- data.frame(
  cell_types = names(jaccard_receptors_receiver),
  jaccard_index = unlist(jaccard_receptors_receiver),
  stringsAsFactors = FALSE
) %>%
  separate(cell_types,
    into = c(
      "sender",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

rownames(jaccard_receptors) <- NULL

# targets
jaccard_targets <- data.frame(
  cell_types = names(jaccard_targets_receiver),
  jaccard_index = unlist(jaccard_targets_receiver),
  stringsAsFactors = FALSE
) %>%
  separate(cell_types,
    into = c(
      "sender",
      "celltype1",
      "celltype2"
    ),
    sep = "_"
  )

rownames(jaccard_targets) <- NULL

# combined
jaccard_receptors$type <- "Receptors"
jaccard_targets$type <- "Targets"

jaccard_all <- rbind(jaccard_receptors, jaccard_targets)
```

# Plotting
* Lollipop plots of JI between excitatory and inhibitory neurons for receptors and targets
* Combined plot of the above
```{r}
png(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "receptors_lollipop_neurons.png"
))
jaccard_receptors %>%
  arrange(jaccard_index) %>%
  mutate(sender = factor(sender, levels = sender)) %>%
  ggplot(aes(x = sender, y = jaccard_index)) +
  geom_segment(aes(xend = sender, yend = 0)) +
  geom_point(size = 4, color = "#6a3d9a") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_blank()
  ) +
  xlab("Sender Cells") +
  ylab("JI") +
  ylim(0, 0.5) +
  ggtitle("Receptors")
dev.off()

png(here(
  "results",
  "intermediate_outputs",
  "04_ccc",
  "targets_lollipop_neurons.png"
))
jaccard_targets %>%
  arrange(jaccard_index) %>%
  mutate(sender = factor(sender, levels = sender)) %>%
  ggplot(aes(x = sender, y = jaccard_index)) +
  geom_segment(aes(xend = sender, yend = 0)) +
  geom_point(size = 4, color = "#cab2d6") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_blank()
  ) +
  xlab("Sender Cells") +
  ylab("JI") +
  ylim(0, 0.5) +
  ggtitle("Targets")
dev.off()

png(here("results", "final_outputs", "03_ccc", "combined_lollipop_neurons.png"))
jaccard_all %>%
  ggplot() +
  geom_linerange(aes(x = sender, ymin = 0, ymax = jaccard_index, colour = type),
    position = position_dodge(width = 0.2), size = 1
  ) +
  geom_point(aes(x = sender, y = jaccard_index, colour = type),
    position = position_dodge(width = 0.2), size = 2
  ) +
  coord_flip() +
  scale_color_manual(
    values = c("Receptors" = "darkgreen", "Targets" = "darkseagreen"),
    name = "Type"
  ) +
  theme_bw() +
  labs(x = "Sender", y = "JI") +
  ylim(0, 0.5) +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  ) +
  ggtitle("JI between Excitatory and Inhibitory Neurons")
dev.off()
```

# Saving outputs
```{r}
saveRDS(
  jaccard_results_ligands,
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "jaccard_ligands_senders.rds"
  )
)

saveRDS(
  jaccard_results_receptors,
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "jaccard_receptors_senders.rds"
  )
)

saveRDS(
  jaccard_results_targets,
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "jaccard_targets_senders.rds"
  )
)

saveRDS(
  jaccard_receptors,
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "jaccard_receptors_receivers.rds"
  )
)

saveRDS(
  jaccard_targets,
  here(
    "results",
    "intermediate_outputs",
    "04_ccc",
    "jaccard_targets_receivers.rds"
  )
)
```

# Session info
```{r}
sessionInfo() # see output below
```
R version 4.2.3 (2023-03-15)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] lintr_3.0.2     styler_1.9.1    here_1.0.1      lubridate_1.9.2 forcats_1.0.0  
 [6] stringr_1.5.0   dplyr_1.1.1     purrr_1.0.1     readr_2.1.4     tidyr_1.3.0    
[11] tibble_3.2.1    ggplot2_3.4.2   tidyverse_2.0.0

loaded via a namespace (and not attached):
 [1] tidyselect_1.2.0  xfun_0.38         remotes_2.4.2     colorspace_2.1-0 
 [5] vctrs_0.6.2       generics_0.1.3    htmltools_0.5.5   yaml_2.3.7       
 [9] utf8_1.2.3        rlang_1.1.0       R.oo_1.25.0       pillar_1.9.0     
[13] glue_1.6.2        withr_2.5.0       R.utils_2.12.2    lifecycle_1.0.3  
[17] R.cache_0.16.0    munsell_0.5.0     gtable_0.3.3      R.methodsS3_1.8.2
[21] evaluate_0.20     labeling_0.4.2    knitr_1.42        tzdb_0.3.0       
[25] callr_3.7.3       fastmap_1.1.1     ps_1.7.5          rex_1.2.1        
[29] fansi_1.0.4       scales_1.2.1      desc_1.4.2        farver_2.1.1     
[33] hms_1.1.3         digest_0.6.31     stringi_1.7.12    processx_3.8.1   
[37] grid_4.2.3        rprojroot_2.0.3   cli_3.6.1         tools_4.2.3      
[41] magrittr_2.0.3    lazyeval_0.2.2    crayon_1.5.2      pkgconfig_2.0.3  
[45] xml2_1.3.3        cyclocomp_1.1.0   timechange_0.2.0  rmarkdown_2.21   
[49] rstudioapi_0.14   R6_2.5.1          compiler_4.2.3

# Time tracking
```{r}
fptm <- proc.time() - ptm

# convert to minutes
(fptm[3] / 60)

# Time elapsed: < 5 minutes
```

# Reproducibility
```{r}
# styler
style_file(here(
  "src",
  "ccc",
  "03_jaccard_similarity.Rmd"
))

# lintr
lint(filename = here(
  "src",
  "ccc",
  "03_jaccard_similarity.Rmd"
))
```
